---
title: "NSPIRE Inspection Optimization Model"
author: "Joël R. Pradines, Ph.D."
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
    math: true
    embed-resources: true
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

devtools::load_all(".")
library(ggplot2)
library(ggridges)
library(viridis)
library(here)
library(tidyr)
library(scales)
library(patchwork)
```

# 1. Introduction

This document outlines the foundational mathematical model for optimizing maintenance resource allocation ahead of an NSPIRE (National Standards for the Physical Inspection of Real Estate) inspection. The objective is to maximize the expected effective score by traversing a dependency-constrained network of tasks, evaluated via Monte Carlo simulations to account for operational stochastics.

# 2. The NSPIRE Scoring Mechanism

The inspection divides a property into distinct locations and evaluates defects based on severity. 

## 2.1. Definitions and Sets
* **Severity:** $s \in \{\text{LT}, \text{S}, \text{STD}\}$ (Life-Threatening, Severe, Standard)
* **Location:** $l \in \{\text{Unit}, \text{Inside}, \text{Outside}\}$
* $N$: Total number of units in the property.
* $n(N)$: The HUD-mandated sample size function, where $n$ is a discrete function of $N$.
* $C_{s,l}$: True total count of defects existing in the property.
* $c_{s,l}$: Observed count of defects recorded during the inspection.
* $w_{s,l}$: Static point weight (penalty) assigned to a defect.

## 2.2. The Deduction Function
The total point deduction $D$ is a summation over all severities and locations. For Unit locations, the observed defects are normalized by the sample size $n(N)$, which effectively scales the deduction based on the inspection tier:

$$D = \sum_{s} \sum_{l \in \{\text{Inside}, \text{Outside}\}} (c_{s,l} \cdot w_{s,l}) + \sum_{s} \left( \frac{c_{s,\text{Unit}}}{n(N)} \cdot w_{s,\text{Unit}} \right)$$

## 2.3. The Raw Continuous Score
The base score originates at 100 and is bounded at zero using the Heaviside step function $H(x)$:

$$S_{raw} = (100 - D) \cdot H(100 - D)$$

## 2.4. The Operational Compliance Function (Failsafe)
A score $\ge 60$ mathematically passes, but the presence of *any* Life-Threatening (LT) defect forces an immediate 24-hour mitigation compliance failure. Let $n_{LT}$ represent the total observed LT defects:

$$n_{LT} = \sum_{l} c_{\text{LT},l}$$

$$S_{eff} = S_{raw} \cdot H(0.5 - n_{LT})$$

# 3. Probabilistic Sampling and Realizations

The inspection score is not a deterministic value but a random variable $S(\omega)$ dependent on the state $\omega$ (the realization of all random variables). 

## 3.1. Stochastic Unit Sampling
HUD inspection protocols dictate that units are sampled without replacement. For a property with $N$ total units, the inspector chooses a subset $n(N)$ based on the tiered sampling schedule defined in Table 1.

| Total Units ($N$) | Inspected Units ($n$) |
| :--- | :--- |
| 1–5 | All units |
| 6–7 | 6 |
| 8 | 7 |
| 9–10 | 8 |
| 11–12 | 9 |
| 13–14 | 10 |
| 15–16 | 11 |
| 17–18 | 12 |
| 19–21 | 13 |
| 22–24 | 14 |
| 25–27 | 15 |
| 28–30 | 16 |
| 31–35 | 17 |
| 36–39 | 18 |
| 40–45 | 19 |
| 46–51 | 20 |
| 52–59 | 21 |
| 60–67 | 22 |
| 68–78 | 23 |
| 79–92 | 24 |
| 93–110 | 25 |
| 111+ | 32 |

While the law of large numbers smooths score variance in high-$N$ properties, this sampling process introduces non-negligible variance in mid-sized properties. Because the inspector's subset is stochastic, the recorded defect count $c_{s,\text{Unit}}$ deviates from the true property state $C_{s,\text{Unit}}$, contributing to the overall spread of the final score distribution.

## 3.2. Realization of Deductions
Let $I_{u}$ be an indicator variable where $I_u = 1$ if unit $u$ is selected for inspection, and $0$ otherwise. The observed defect count $c_{s,\text{Unit}}$ for a specific simulation realization $\omega$ is defined by the subset of selected units:

$$c_{s,\text{Unit}}(\omega) = \sum_{u=1}^{N} I_u(\omega) \cdot \mathbb{1}(\text{unit } u \text{ has defect } s)$$

Unlike an expected value approach, our model calculates the realization $D(\omega)$ for each iteration $k$ in the Monte Carlo ensemble. This preserves the variance inherent in the sampling process, which is the primary driver of the "lottery effect" in mid-sized properties.

# 4. Resource Mapping and Graph Topology (DAG)

Maintenance tasks form a **Directed Acyclic Graph (DAG)** where each task $i$ is a vertex $V_i$ with an expected execution time $t_i$ and material cost $m_i$.

## 4.1. Edge Constraints
1.  **Sequential Precedence:** A directed edge $V_A \to V_B$ requires the start time of $B$ to be $\ge$ the end time of $A$.
2.  **Temporal Blocking:** Certain vertices impose a local resource freeze, locking a spatial node for a time interval $\Delta t$.
3.  **Mutually Exclusive Pruning:** If a higher-order vertex $V_{replace}$ is activated, redundant child vertices $V_{repair}$ are pruned ($t_{repair} \to 0$).

# 5. Task Modalities & Execution Constraints

Task execution times are not normally distributed; they represent distinct physical realities (e.g., a quick adjustment versus a catastrophic replacement). 

## 5.1. Mixture Distributions (PERT)
Task $i$ exists in state $k \in \{1, 2, 3\}$. The execution time $t_i$ is drawn from a bounded Beta-PERT distribution, weighted by the physical probability of the state ($p_{i,k}$):

$$f(t_i) = \sum_{k} p_{i,k} \cdot \text{PERT}(t; a_k, m_k, b_k)$$
where $a_k, m_k, b_k$ represent the optimistic, mode, and pessimistic completion times.

## 5.2. Vendor Handoff Nodes
If a task evaluates to a severe state requiring external licensing or specialization (e.g., licensed plumbing or glazing), the node logic bifurcates:
* Internal labor time drops to zero ($t_{internal} \to 0$).
* A schedule lag is injected, drawing from a Uniform distribution: $L \sim \text{Uniform}(l_{min}, l_{max})$ days.
* A mandatory spatial lockout is triggered (e.g., 24-hour cure time for silicone), forcing the DAG to re-route internal technicians.

# 6. Quality Control & Redundancy

Execution of a task does not guarantee absolute resolution. A task has a baseline probability ($p_{err}$) of being missed or improperly repaired.

## 6.1. Audit Efficacy
We introduce a formal internal audit efficacy parameter, $\alpha_\text{audit} \in [0, 1]$, representing the probability that the pre-inspection internal team detects and successfully mitigates a defect. 

If a defect survives the initial repair attempt (with probability $p_{err}$), the probability that it remains undetected by the audit phase is:
$$P(\text{Survive}) = (1 - \alpha_\text{audit}) \cdot p_{err}$$

## 6.2. Technician Redundancy
* **Serial Execution (Single Technician):** The probability of a defect surviving into the inspection phase follows the logic in 6.1.
* **Parallel Execution (Dual Technicians):** Applying the "Four Eyes" principle, independent cross-checking drastically reduces error. The joint probability of both technicians missing the defect becomes $P(\text{Survive}) = (1 - \alpha_\text{audit}) \cdot p_{err}^2$.

# 7. Temporal Windowing and Asset Decay

The operational model assumes a maintenance repair window defined by $t_{start}$ (commencement of internal repairs) and $t_{end}$ (completion of repairs/audit). The interval $\delta = t_{start} - t_{end}$ constrains the total available labor capacity, while the subsequent interval $\delta_{decay} = t_{inspection} - t_{end}$ dictates the post-repair asset degradation.

## 7.1. Repair Capacity
Given a standard base labor capacity $R_{base}$ (hours/unit/day), the total labor capacity $\text{Cap}$ available for the maintenance cycle is linearly proportional to the duration of the repair window:


$$\text{Cap} = R_{base} \cdot (t_{start} - t_{end})$$

Tasks requiring labor hours $T_{total} > \text{Cap}$ result in a scheduling backlog, forcing the prioritization of high-penalty items (LT/S) at the expense of lower-severity items.

## 7.2. Tenant Heterogeneity (Bimodal Risk)
Asset deterioration is not uniform across a property. We model unit "cleanliness" as a bimodal latent state $\phi_u$ for each unit $u$, representing the inherent maintenance burden generated by the tenant. The tenant impact multiplier $\phi_u$ is drawn from a mixture distribution:

$$\phi_u \sim \begin{cases} \phi_{clean}, & \text{with probability } (1 - \pi) \\ \phi_{dirty}, & \text{with probability } \pi \end{cases}$$

Where $\pi$ is the estimated proportion of high-impact units in the property.

## 7.3. Post-Audit Asset Decay
Once repairs are audited at $t_{end}$, the asset re-enters a state of stochastic wear. We model the defect re-emergence as a Bernoulli process over the decay window $\delta_{decay}$. 

Given a daily probability of failure $\lambda_{i}$ for task $i$, the probability $P_{redefect}$ that a resolved task fails before the inspection date is:

$$P_{redefect} = 1 - (1 - \lambda_{i,u}^*)^{\delta_{decay}}$$

The effective daily probability of failure $\lambda_{i,u}^*$ is coupled with the unit's latent state $\phi_u$, effectively accelerating the re-emergence of defects in high-impact ("dirty") units:

$$\lambda_{i,u}^* = \lambda_{i} \cdot \phi_u$$

This temporal decay creates a trade-off: an earlier $t_{end}$ increases the decay window, inflating the probability that fixed items will fail before the inspector arrives, while a late $t_{end}$ constricts the repair window, potentially inducing a labor backlog.

# 8. Predictive Topological Risk (Latent Covariates)

Baseline task probabilities assume a standard asset deterioration rate. However, the lack of historical tenant reporting often masks hidden deferred maintenance, warping the DAG's risk topology. We use historical service request volume $W_u$ as a proxy to estimate the probability that a unit belongs to the "dirty" latent state ($\phi_{dirty}$).

## 8.1. Inferring Latent States
We define the unit-specific probability $\pi_u$ of a unit being in the high-impact state as a logistic function of service requests $W_u$:

$$\pi_u = P(\phi_u = \phi_{dirty} | W_u) = \text{logit}^{-1}(\alpha - \beta W_u)$$

Where $\alpha$ and $\beta$ are calibrated tuning parameters. For units with low reporting ($W_u \to 0$), the prior probability of being in the "dirty" state increases, signaling a higher likelihood of latent deferred maintenance.

## 8.2. Dynamic Risk Mapping

This unit-specific $\pi_u$ dynamically replaces the global mixture probability used in Section 7.2. During simulation initialization, each unit's tenant factor $\phi_u$ is sampled conditioned on its specific reporting history:

$$\phi_u \sim \begin{cases} \phi_{clean}, & \text{with probability } (1 - \pi_u) \\ \phi_{dirty}, & \text{with probability } \pi_u \end{cases}$$

This approach transforms the simulation from a static property model into a predictive prioritization tool. By mapping historical service request volumes to the bimodal risk topology, the engine directs the maintenance DAG to prioritize units where $W_u$ is low (indicating a high probability of "hidden" defects) before they enter the randomized HUD sampling window.

```{r}
#| label: tbl-tasks-config
#| tbl-cap: "NSPIRE Task Modalities and Execution Constraints"
#| echo: false
#| warning: false
#| message: false

# Load the required libraries
library(DT)
library(readr)

# Read the CSV
tasks_df <- read_csv("../inst/extdata/tasks_config.csv")

# Generate and print the interactive table
datatable(
  tasks_df,
  extensions = c('FixedColumns', 'Responsive'),
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    fixedColumns = list(leftColumns = 5), 
    autoWidth = TRUE,
    columnDefs = list(
      list(className = 'dt-center', targets = "_all")
    )
  ),
  class = 'cell-border stripe hover'
)
```

# 9. Risk Analysis

Moving beyond deterministic score projections, this model quantifies operational risk by treating property performance as a stochastic process. By leveraging Monte Carlo simulation, we assess not only the expected outcome but the probability of regulatory non-compliance.

## 9.1. Quantification of Failure Risk

We define the failure event as $S_{eff} < 60$. Due to the non-linear interaction of random unit sampling, defect emergence probabilities, and repair quality, the inspection score $S_{eff}$ is a random variable. We estimate the probability of failure ($P_{fail}$) by simulating $K = 1,000$ independent inspection realizations:

$$P_{fail} = \frac{1}{K} \sum_{i=1}^{K} I(S_{eff,i} < 60)$$

where $I(\cdot)$ is the indicator function. This allows us to map the property's risk profile against the HUD enforcement threshold, providing a granular view of the likelihood of catastrophic failure that a mean-based score effectively masks.

## 9.2. Quantifying Volatility: The Tail Drop Metric

Standard deviation ($\sigma$) and the Coefficient of Variation ($CV = \sigma/\mu$) provide a global measure of dispersion. However, in NSPIRE compliance, the right tail (scores $> 100$) is truncated, and the distribution is often skewed. The primary operational risk resides in the left tail—the possibility of a "bad day" where sampling captures a cluster of high-severity defects.

To quantify this downside volatility, we define the **Tail Drop Metric** ($\tau$):

$$\tau = \frac{\text{med} - P_{05}}{\text{med}}$$ {#eq-tau}

Where:

*   $\text{med}$ is the median score.
*   $P_{05}$ is the 5th percentile of the simulated scores (the "Value at Risk").

$\tau$ represents the percentage degradation from expected performance to the worst-case scenario. A property may exhibit a high $\text{med}$ while maintaining a high $\tau$, identifying it as a "lottery" asset—stable on average, but highly susceptible to non-linear drops in score due to the interaction of unit sampling variance and tenant-driven decay.

## 9.3. Sensitivity Mapping

The Tail Drop Metric $\tau$ is a systemic output of our model, functionally defined by the interaction of our control and stochastic parameters:
$$\tau = f(n(N), \alpha_\text{audit}, \pi, \phi_{dirty})$$

This sensitivity function captures how variance in the HUD sampling tier $n(N)$, human error in pre-inspection audit ($\alpha_\text{audit}$), and the underlying distribution of tenant impact ($\pi, \phi_{dirty}$) propagate into downside performance risk. Future numerical investigations will isolate these partial derivatives to determine which parameter dominates the variance in the $P_{05}$ left tail.

---

# 10. Parameter Sensitivity Analysis

Sensitivity analysis studies how variations in control parameters drive system outcomes, such as expected score, labor costs, and downside risk ($P_{fail}$ and $\tau$). By studying these variables across parameter ranges, one can identify parameter critical ranges, non-linear thresholds and propose optimal resource allocations within the NSPIRE framework.

## 10.1. Audit Catch Rate ($\alpha_\text{audit}$)

The internal pre-inspection audit is modeled by $\alpha_\text{audit}$, the probability that a defect is identified and resolved prior to the HUD inspection. To quantify the impact of audit efficacy on compliance risk, $\alpha_\text{audit}$ was varied from $0.05$ (negligent) to $0.95$ (highly rigorous), while executing $K=1,000$ Monte Carlo iterations per step. @fig-audit-ridges provides a visual summary of the resulting score distributions (horizontal axis with a red line for the failing score of 60) as a function of $\alpha_\text{audit}$ (vertical axis).

```{r}
#| label: fig-audit-ridges
#| fig-width: 9
#| fig-height: 6
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Risk Transition: Audit Rate vs. Score Distribution."

sweep_data <- load_sensitivity_iterations(here("results", "sensitivity", "audit_rate"))

# Generate the plot
plot_sensitivity_ridges(
  sensitivity_df = sweep_data,
  param_label = "Audit Catch Rate (Efficacy)",
  title = "Audit Rate vs. Score Distribution"
)
```

Failure risk is almost completely eliminated when $\alpha_\text{audit} \geq 0.9$. @fig-distribution-slice further explains the transition from high risk to low risk as $\alpha_\text{audit}$ increases from 0.55 to 0.85. The failure risk is best estimated with the proportion of simulations (among $K=1,000$) that yield a score below 60 (red vertical line). Visually, this corresponds to the area under the density curve (integral) to the left of the critical red line. 


```{r} 
#| label: fig-distribution-slice
#| fig.width: 12 
#| fig.height: 6
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Examples of Score Distribution: $\\alpha_\\text{audit} = 0.55$ (left) and $\\alpha_\\text{audit} = 0.85$ (right)."

# 1. Define the scenarios
scenarios <- c(0.55, 0.85)
sum_df <- readRDS(here("results", "sensitivity", "audit_rate", "summary.rds"))
pal <- scales::viridis_pal(option = "D")(100) # Pre-generate palette

plots <- lapply(scenarios, function(rate) {
  # 1. Cast both to numeric to ensure they match
  median_val <- sum_df$score_median[as.numeric(sum_df$param_val) == as.numeric(rate)]
  
  # 2. Pick color based on median score (0-100 mapped to 1-100 index)
  idx <- max(1, min(100, round(median_val)))
  
  plot_score_distribution(
    input = here("results", "sensitivity", "audit_rate", paste0("score_density_", rate, ".rds")),
    title = paste("Audit Catch Rate:", rate * 100, "%"),
    color_hex = pal[idx]
  )
})

plots[[1]] | plots[[2]]

```

Estimating both failure risks (score less than 30 or less than 60) at all values of $\alpha_\text{audit}$ yields @fig-failure-audit. The data demonstrates that critical failure risk (score < 30) is only marginalized when audit catch rates exceed 0.75, while full protection against standard failure (score < 60) requires efficacy above 0.90.

```{r}
#| label: fig-failure-audit
#| fig.width: 6
#| fig.height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Failure probabilities at scores 30 and 60 as a function of audit catch rate $\\alpha_\\text{audit}$."

sum_df <- readRDS(here("results", "sensitivity", "audit_rate", "summary.rds"))

plot_sensitivity_multi(
  sum_df, 
  y_vars = c("fail_risk_60", "critical_fail_risk_30"),
  param_label = "Audit Catch Rate",
  y_label = "Probability",
  title = "Impact of Audit Efficacy on Regulatory Failure"
)
```

Besides failure probabilities, another useful statistic is the tail-drop metric $\tau$ @eq-tau. To understand its sensitivity to $\alpha_\text{audit}$, we must first examine how the foundational score statistics evolve across the parameter range. These plots are displayed in @fig-dist-stats-audit.

```{r}
#| label: fig-dist-stats-audit
#| fig.width: 6
#| fig.height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Score statistics as a function of audit catch rate $\\alpha_\\text{audit}$."

sum_df <- readRDS(here("results", "sensitivity", "audit_rate", "summary.rds"))

plot_sensitivity_multi(
  sum_df, 
  y_vars = c("score_p05", "score_median", "score_sd"),
  param_label = "Audit Catch Rate",
  y_label = "Score",
  title = "Impact of Audit Efficacy on Score Statistics"
)
```

As $\alpha_\text{audit}$ increases beyond 0.25, the score median starts rising linearly. Later, as $\alpha_\text{audit}$ gets above 0.45, the 5th percentile score $P_{05}$ also rises linearly but with a stiffer slope. This yields decreasing differences between median and 5th percentile. One way to visualize this is to plot the standard deviation of the score (a proxy measure for difference between median and 5th percentile) as a function of $\alpha_\text{audit}$. These behaviors of the different statistics yield the profile of $\tau$ @eq-tau as a function of $\alpha_\text{audit}$ displayed in @fig-tau-audit.

```{r}
#| label: fig-tau-audit
#| fig.width: 6
#| fig.height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Tail drop metric $\\tau$ as a function of audit catch rate $\\alpha_\\text{audit}$."

sum_df <- readRDS(here("results", "sensitivity", "audit_rate", "summary.rds"))

plot_sensitivity_multi(
  sum_df, 
  y_vars = c("tail_drop_median"),
  param_label = "Audit Catch Rate",
  y_label = "Tail Drop Metric",
  title = "Impact of Audit Efficacy on Tail Drop Metric."
)
```

**Conclusions for $\alpha_\text{audit}$**:

* Operational Futility Threshold ($\alpha_\text{audit} \leq 0.25$): Below ~25%
efficacy, quality assurance efforts are mathematically invisible to the final metric due to the absolute scoring floor (0).

* Variance Squeeze ($0.45 < \alpha_\text{audit} < 0.90$): As the catch rate rises above 0.45, the 5th percentile worst-case outcome ($P_{05}$) improves at a faster linear rate than the median. This compresses the score variance and significantly reduces the Tail Drop metric ($\tau$).

* Safety Baseline ($\alpha_\text{audit} \geq 0.90$): To confidently eliminate standard regulatory failure risk (score < 60), the internal audit process must successfully intercept at least 90% of all pre-existing defects.

---

## 10.2. The Time Crunch: Notice of Inspection ($t_\text{start}$)

While a robust audit catch rate ($\alpha_\text{audit}$) is critical for passing an NSPIRE inspection, it assumes the maintenance team has the physical time to execute the repairs. HUD inspection notices can vary drastically, sometimes providing as little as 14 to 30 days of warning.

To model this operational squeeze, the preparation window ($t_\text{start}$) was swept from 36 days down to 4 days. For this specific scenario, the internal audit catch rate was fixed at a highly effective 85% ($\alpha_\text{audit} = 0.85$), and maximum allowable overtime was expanded to 4 hours per unit per day. Crucially, the simulation incorporates context-aware transit friction: technicians incur small delays (1–5 minutes) when moving between tasks in the same unit, but significant delays (15–30 minutes) when packing up and accessing a new unit.

As the preparation window shrinks, the volume of identified defects remains constant, but the time available to clear them—and the time available to walk between them—compresses. The initial impact is purely financial, as visualized in @fig-time-crunch-labor.

```{r}
#| label: fig-time-crunch-labor
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Mean total labor cost and overtime cost as the preparation window ($t_\\text{start}$) changes. Note the massive overtime spike at Day 22 followed by the sharp drop in total costs as the timeline collapses."

sum_df_time <- readRDS(here("results", "sensitivity", "t_start", "master_time_crunch_study.rds"))

plot_sensitivity_multi(
  sum_df_time, 
  y_vars = c("total_cost_mean", "overtime_mean"),
  param_label = "Days of Notice (t_start)",
  y_label = "Cost ($)",
  title = "Financial Impact of Compressed Prep Windows"
)
```

At first glance, the left side of @fig-time-crunch-labor appears counter-intuitive: as the preparation window drops below 20 days, the total expected cost plunges dramatically. This is not an efficiency gain; it is the mathematical manifestation of deferred maintenance.

Between Day 30 and Day 22, the workforce heavily leverages their overtime allowance to keep pace with the workload. At Day 22, the system hits peak labor output—the team is operating at absolute maximum capacity, driving overtime spend to its highest point. However, as the window shrinks below 20 days, the absolute physical capacity of the workforce is exhausted. Hundreds of items are forced into the "backlog." Because backlogged items are abandoned, they incur no labor and no material costs, creating a "false economy" of savings.

@fig-time-crunch-backlog illustrates the severe regulatory consequences of this capacity exhaustion.

```{r}
#| label: fig-time-crunch-backlog
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Probability of generating a task backlog and the resulting regulatory failure risk ($<60$) under compressed timelines."

plot_sensitivity_multi(
  sum_df_time, 
  y_vars = c("backlog_prob", "fail_risk_60"),
  param_label = "Days of Notice (t_start)",
  y_label = "Probability",
  title = "The Backlog Cliff: Capacity Exhaustion"
)
```

**Conclusions for $t_\text{start}$**:

**Distributed Defect Tax ($t_\text{start} \geq 28$)**: Because repairing widely distributed defects incurs a heavy transit-time penalty, a 28-day notice is the edge of the safe operational limit for this property. At 28 days, the property maintains a near 0% failure risk with minimal overtime expenditure.

**Maximum Effort Peak ($t_\text{start} \approx 22$)**: Day 22 represents the system's breaking point. Overtime spend reaches its absolute maximum as the team attempts to outwork the shrinking timeline. Despite spending heavily on premium labor, the backlog probability reaches ~50%, and the property begins to exhibit a measurable risk of regulatory failure. The team is throwing maximum capital at the problem but is still losing ground.

**Operational Collapse ($t_\text{start} \leq 15$)**: A standard 14-day HUD notice is mathematically unfeasible for a property in this state and with 5 technicians. By Day 15, the probability of backlogging tasks reaches 100%, and the property's risk of failure exceeds 85%. The simulation proves that sudden drops in preparation expenditure in the days leading up to an inspection are a lagging indicator of capacity exhaustion, representing a forced trade-off of immediate OpEx savings for guaranteed regulatory failure.

---

<style>
.floating-copyright {
  position: fixed;
  bottom: 15px;
  left: 15px;
  font-size: 0.85em;
  color: #777;
  background-color: rgba(255, 255, 255, 0.85);
  padding: 5px 10px;
  border-radius: 5px;
  z-index: 1000;
  pointer-events: none; 
}
</style>

<div class="floating-copyright">
  &copy; 2026 Joël R. Pradines.<br> All rights reserved.
</div>