---
title: "NSPIRE Inspection Optimization Model"
author: "Joël R. Pradines, Ph.D."
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
    math: true
    embed-resources: true
---

```{r}
devtools::load_all(".")

# 2. Load the external libraries needed for the joyplot
library(ggplot2)
library(ggridges)
library(viridis)
library(here)
library(tidyr)
```

# 1. Introduction

This document outlines the foundational mathematical model for optimizing maintenance resource allocation ahead of an NSPIRE (National Standards for the Physical Inspection of Real Estate) inspection. The objective is to maximize the expected effective score by traversing a dependency-constrained network of tasks, evaluated via Monte Carlo simulations to account for operational stochastics.

# 2. The NSPIRE Scoring Mechanism

The inspection divides a property into distinct locations and evaluates defects based on severity. 

## 2.1. Definitions and Sets
* **Severity:** $s \in \{\text{LT}, \text{S}, \text{STD}\}$ (Life-Threatening, Severe, Standard)
* **Location:** $l \in \{\text{Unit}, \text{Inside}, \text{Outside}\}$
* $N$: Total number of units in the property.
* $n(N)$: The HUD-mandated sample size function, where $n$ is a discrete function of $N$.
* $C_{s,l}$: True total count of defects existing in the property.
* $c_{s,l}$: Observed count of defects recorded during the inspection.
* $w_{s,l}$: Static point weight (penalty) assigned to a defect.

## 2.2. The Deduction Function
The total point deduction $D$ is a summation over all severities and locations. For Unit locations, the observed defects are normalized by the sample size $n(N)$, which effectively scales the deduction based on the inspection tier:

$$D = \sum_{s} \sum_{l \in \{\text{Inside}, \text{Outside}\}} (c_{s,l} \cdot w_{s,l}) + \sum_{s} \left( \frac{c_{s,\text{Unit}}}{n(N)} \cdot w_{s,\text{Unit}} \right)$$

## 2.3. The Raw Continuous Score
The base score originates at 100 and is bounded at zero using the Heaviside step function $H(x)$:

$$S_{raw} = (100 - D) \cdot H(100 - D)$$

## 2.4. The Operational Compliance Function (Failsafe)
A score $\ge 60$ mathematically passes, but the presence of *any* Life-Threatening (LT) defect forces an immediate 24-hour mitigation compliance failure. Let $n_{LT}$ represent the total observed LT defects:

$$n_{LT} = \sum_{l} c_{\text{LT},l}$$

$$S_{eff} = S_{raw} \cdot H(0.5 - n_{LT})$$

# 3. Probabilistic Sampling and Realizations

The inspection score is not a deterministic value but a random variable $S(\omega)$ dependent on the state $\omega$ (the realization of all random variables). 

## 3.1. Stochastic Unit Sampling
HUD inspection protocols dictate that units are sampled without replacement. For a property with $N$ total units, the inspector chooses a subset $n(N)$ based on the tiered sampling schedule defined in Table 1.

| Total Units ($N$) | Inspected Units ($n$) |
| :--- | :--- |
| 1–5 | All units |
| 6–7 | 6 |
| 8 | 7 |
| 9–10 | 8 |
| 11–12 | 9 |
| 13–14 | 10 |
| 15–16 | 11 |
| 17–18 | 12 |
| 19–21 | 13 |
| 22–24 | 14 |
| 25–27 | 15 |
| 28–30 | 16 |
| 31–35 | 17 |
| 36–39 | 18 |
| 40–45 | 19 |
| 46–51 | 20 |
| 52–59 | 21 |
| 60–67 | 22 |
| 68–78 | 23 |
| 79–92 | 24 |
| 93–110 | 25 |
| 111+ | 32 |

While the law of large numbers smooths score variance in high-$N$ properties, this sampling process introduces non-negligible variance in mid-sized properties. Because the inspector's subset is stochastic, the recorded defect count $c_{s,\text{Unit}}$ deviates from the true property state $C_{s,\text{Unit}}$, contributing to the overall spread of the final score distribution.

## 3.2. Realization of Deductions
Let $I_{u}$ be an indicator variable where $I_u = 1$ if unit $u$ is selected for inspection, and $0$ otherwise. The observed defect count $c_{s,\text{Unit}}$ for a specific simulation realization $\omega$ is defined by the subset of selected units:

$$c_{s,\text{Unit}}(\omega) = \sum_{u=1}^{N} I_u(\omega) \cdot \mathbb{1}(\text{unit } u \text{ has defect } s)$$

Unlike an expected value approach, our model calculates the realization $D(\omega)$ for each iteration $k$ in the Monte Carlo ensemble. This preserves the variance inherent in the sampling process, which is the primary driver of the "lottery effect" in mid-sized properties.

# 4. Resource Mapping and Graph Topology (DAG)

Maintenance tasks form a **Directed Acyclic Graph (DAG)** where each task $i$ is a vertex $V_i$ with an expected execution time $t_i$ and material cost $m_i$.

## 4.1. Edge Constraints
1.  **Sequential Precedence:** A directed edge $V_A \to V_B$ requires the start time of $B$ to be $\ge$ the end time of $A$.
2.  **Temporal Blocking:** Certain vertices impose a local resource freeze, locking a spatial node for a time interval $\Delta t$.
3.  **Mutually Exclusive Pruning:** If a higher-order vertex $V_{replace}$ is activated, redundant child vertices $V_{repair}$ are pruned ($t_{repair} \to 0$).

# 5. Task Modalities & Execution Constraints

Task execution times are not normally distributed; they represent distinct physical realities (e.g., a quick adjustment versus a catastrophic replacement). 

## 5.1. Mixture Distributions (PERT)
Task $i$ exists in state $k \in \{1, 2, 3\}$. The execution time $t_i$ is drawn from a bounded Beta-PERT distribution, weighted by the physical probability of the state ($p_{i,k}$):

$$f(t_i) = \sum_{k} p_{i,k} \cdot \text{PERT}(t; a_k, m_k, b_k)$$
where $a_k, m_k, b_k$ represent the optimistic, mode, and pessimistic completion times.

## 5.2. Vendor Handoff Nodes
If a task evaluates to a severe state requiring external licensing or specialization (e.g., licensed plumbing or glazing), the node logic bifurcates:
* Internal labor time drops to zero ($t_{internal} \to 0$).
* A schedule lag is injected, drawing from a Uniform distribution: $L \sim \text{Uniform}(l_{min}, l_{max})$ days.
* A mandatory spatial lockout is triggered (e.g., 24-hour cure time for silicone), forcing the DAG to re-route internal technicians.

# 6. Quality Control & Redundancy

Execution of a task does not guarantee absolute resolution. A task has a baseline probability ($p_{err}$) of being missed or improperly repaired.

## 6.1. Audit Efficacy
We introduce a formal internal audit efficacy parameter, $\alpha_{audit} \in [0, 1]$, representing the probability that the pre-inspection internal team detects and successfully mitigates a defect. 

If a defect survives the initial repair attempt (with probability $p_{err}$), the probability that it remains undetected by the audit phase is:
$$P(\text{Survive}) = (1 - \alpha_{audit}) \cdot p_{err}$$

## 6.2. Technician Redundancy
* **Serial Execution (Single Technician):** The probability of a defect surviving into the inspection phase follows the logic in 6.1.
* **Parallel Execution (Dual Technicians):** Applying the "Four Eyes" principle, independent cross-checking drastically reduces error. The joint probability of both technicians missing the defect becomes $P(\text{Survive}) = (1 - \alpha_{audit}) \cdot p_{err}^2$.

# 7. Temporal Windowing and Asset Decay

The operational model assumes a maintenance repair window defined by $t_{start}$ (commencement of internal repairs) and $t_{end}$ (completion of repairs/audit). The interval $\delta = t_{start} - t_{end}$ constrains the total available labor capacity, while the subsequent interval $\delta_{decay} = t_{inspection} - t_{end}$ dictates the post-repair asset degradation.

## 7.1. Repair Capacity
Given a standard base labor capacity $R_{base}$ (hours/unit/day), the total labor capacity $\text{Cap}$ available for the maintenance cycle is linearly proportional to the duration of the repair window:


$$\text{Cap} = R_{base} \cdot (t_{start} - t_{end})$$

Tasks requiring labor hours $T_{total} > \text{Cap}$ result in a scheduling backlog, forcing the prioritization of high-penalty items (LT/S) at the expense of lower-severity items.

## 7.2. Tenant Heterogeneity (Bimodal Risk)
Asset deterioration is not uniform across a property. We model unit "cleanliness" as a bimodal latent state $\phi_u$ for each unit $u$, representing the inherent maintenance burden generated by the tenant. The tenant impact multiplier $\phi_u$ is drawn from a mixture distribution:

$$\phi_u \sim \begin{cases} \phi_{clean}, & \text{with probability } (1 - \pi) \\ \phi_{dirty}, & \text{with probability } \pi \end{cases}$$

Where $\pi$ is the estimated proportion of high-impact units in the property.

## 7.3. Post-Audit Asset Decay
Once repairs are audited at $t_{end}$, the asset re-enters a state of stochastic wear. We model the defect re-emergence as a Bernoulli process over the decay window $\delta_{decay}$. 

Given a daily probability of failure $\lambda_{i}$ for task $i$, the probability $P_{redefect}$ that a resolved task fails before the inspection date is:

$$P_{redefect} = 1 - (1 - \lambda_{i,u}^*)^{\delta_{decay}}$$

The effective daily probability of failure $\lambda_{i,u}^*$ is coupled with the unit's latent state $\phi_u$, effectively accelerating the re-emergence of defects in high-impact ("dirty") units:

$$\lambda_{i,u}^* = \lambda_{i} \cdot \phi_u$$

This temporal decay creates a trade-off: an earlier $t_{end}$ increases the decay window, inflating the probability that fixed items will fail before the inspector arrives, while a late $t_{end}$ constricts the repair window, potentially inducing a labor backlog.

# 8. Predictive Topological Risk (Latent Covariates)

Baseline task probabilities assume a standard asset deterioration rate. However, the lack of historical tenant reporting often masks hidden deferred maintenance, warping the DAG's risk topology. We use historical service request volume $W_u$ as a proxy to estimate the probability that a unit belongs to the "dirty" latent state ($\phi_{dirty}$).

## 8.1. Inferring Latent States
We define the unit-specific probability $\pi_u$ of a unit being in the high-impact state as a logistic function of service requests $W_u$:

$$\pi_u = P(\phi_u = \phi_{dirty} | W_u) = \text{logit}^{-1}(\alpha - \beta W_u)$$

Where $\alpha$ and $\beta$ are calibrated tuning parameters. For units with low reporting ($W_u \to 0$), the prior probability of being in the "dirty" state increases, signaling a higher likelihood of latent deferred maintenance.

## 8.2. Dynamic Risk Mapping

This unit-specific $\pi_u$ dynamically replaces the global mixture probability used in Section 7.2. During simulation initialization, each unit's tenant factor $\phi_u$ is sampled conditioned on its specific reporting history:

$$\phi_u \sim \begin{cases} \phi_{clean}, & \text{with probability } (1 - \pi_u) \\ \phi_{dirty}, & \text{with probability } \pi_u \end{cases}$$

This approach transforms the simulation from a static property model into a predictive prioritization tool. By mapping historical service request volumes to the bimodal risk topology, the engine directs the maintenance DAG to prioritize units where $W_u$ is low (indicating a high probability of "hidden" defects) before they enter the randomized HUD sampling window.

```{r}
#| label: tbl-tasks-config
#| tbl-cap: "NSPIRE Task Modalities and Execution Constraints"
#| echo: false
#| warning: false
#| message: false

# Load the required libraries
library(DT)
library(readr)

# Read the CSV
tasks_df <- read_csv("../inst/extdata/tasks_config.csv")

# Generate and print the interactive table
datatable(
  tasks_df,
  extensions = c('FixedColumns', 'Responsive'),
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    fixedColumns = list(leftColumns = 5), 
    autoWidth = TRUE,
    columnDefs = list(
      list(className = 'dt-center', targets = "_all")
    )
  ),
  class = 'cell-border stripe hover'
)
```

# 9. Risk Analysis

Moving beyond deterministic score projections, this model quantifies operational risk by treating property performance as a stochastic process. By leveraging Monte Carlo simulation, we assess not only the expected outcome but the probability of regulatory non-compliance.

## 9.1. Quantification of Failure Risk

We define the failure event as $S_{eff} < 60$. Due to the non-linear interaction of random unit sampling, defect emergence probabilities, and repair quality, the inspection score $S_{eff}$ is a random variable. We estimate the probability of failure ($P_{fail}$) by simulating $K = 1,000$ independent inspection realizations:

$$P_{fail} = \frac{1}{K} \sum_{i=1}^{K} I(S_{eff,i} < 60)$$

where $I(\cdot)$ is the indicator function. This allows us to map the property's risk profile against the HUD enforcement threshold, providing a granular view of the likelihood of catastrophic failure that a mean-based score effectively masks.

## 9.2. Quantifying Volatility: The Tail Drop Metric

Standard deviation ($\sigma$) and the Coefficient of Variation ($CV = \sigma/\mu$) provide a global measure of dispersion. However, in NSPIRE compliance, the right tail (scores $> 100$) is truncated, and the distribution is often skewed. The primary operational risk resides in the left tail—the possibility of a "bad day" where sampling captures a cluster of high-severity defects.

To quantify this downside volatility, we define the **Tail Drop Metric** ($\tau$):

$$\tau = \frac{\text{med} - Z_{05}}{\text{med}}$$

Where:

*   $\text{med}$ is the median score.
*   $Z_{05}$ is the 5th percentile of the simulated scores (the "Value at Risk").

$\tau$ represents the percentage degradation from expected performance to the worst-case scenario. A property may exhibit a high $\text{med}$ while maintaining a high $\tau$, identifying it as a "lottery" asset—stable on average, but highly susceptible to non-linear drops in score due to the interaction of unit sampling variance and tenant-driven decay.

## 9.3. Sensitivity Mapping

The Tail Drop Metric $\tau$ is a systemic output of our model, functionally defined by the interaction of our control and stochastic parameters:
$$\tau = f(n(N), \alpha_{audit}, \pi, \phi_{dirty})$$

This sensitivity function captures how variance in the HUD sampling tier $n(N)$, human error in pre-inspection audit ($\alpha_{audit}$), and the underlying distribution of tenant impact ($\pi, \phi_{dirty}$) propagate into downside performance risk. Future numerical investigations will isolate these partial derivatives to determine which parameter dominates the variance in the $Z_{05}$ left tail.

---

# 10. Parameter Sensitivity Analysis

Sensitivity analysis maps how variations in control parameters drive system outcomes, such as expected score, labor costs, and downside risk ($P_{fail}$ and $\tau$). By studying these variables across parameter ranges, we can identify parameter critical ranges, non-linear thresholds and pinpoint optimal resource allocations within the NSPIRE framework.

## 10.1. Audit Catch Rate ($\alpha_{audit}$)

The internal pre-inspection audit is modeled by $\alpha_{audit}$, the probability that a defect is identified and resolved prior to the HUD inspection. To quantify the impact of audit efficacy on compliance risk, we varied $\alpha_{audit}$ from $0.05$ (negligent) to $0.95$ (highly rigorous), executing $K=1,000$ Monte Carlo iterations per step.

```{r}
sweep_data <- load_sensitivity_iterations(here("results", "sensitivity", "audit_rate"))

# Generate the plot
plot_sensitivity_ridges(
  sensitivity_df = sweep_data,
  param_label = "Audit Catch Rate (Efficacy)",
  title = "Systemic Risk Transition: Audit Rate vs. Score Distribution"
)
```

```{r, sensitivity-gallery, fig-width: 6, fig-height: 4}
sum_df <- readRDS(here("results", "sensitivity", "audit_rate", "summary.rds"))

# 1. Failure Risk ($P_{fail}$)
plot_sensitivity_metric(sum_df, "fail_risk_60", "Audit Catch Rate", "Prob. of Failure (Score < 60)", "Systemic Failure Risk")

# 2. Tail Drop Volatility ($\tau$)
plot_sensitivity_metric(sum_df, "tail_drop_median", "Audit Catch Rate", "Tail Drop Metric (τ)", "Downside Volatility")

plot_sensitivity_multi(
  sum_df, 
  y_vars = c("fail_risk_60", "critical_fail_risk_30"),
  param_label = "Audit Catch Rate",
  y_label = "Probability",
  title = "Impact of Audit Efficacy on Regulatory Failure"
)

```

---

<style>
.floating-copyright {
  position: fixed;
  bottom: 15px;
  left: 15px;
  font-size: 0.85em;
  color: #777;
  background-color: rgba(255, 255, 255, 0.85);
  padding: 5px 10px;
  border-radius: 5px;
  z-index: 1000;
  pointer-events: none; 
}
</style>

<div class="floating-copyright">
  &copy; 2026 Joël R. Pradines.<br> All rights reserved.
</div>