---
title: "Optimization of Pre-NSPIRE Inspection Work via Monte Carlo Simulation"
author: "Joël R. Pradines, Ph.D."
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
    math: true
    embed-resources: true
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

devtools::load_all(".")
library(ggplot2)
library(ggridges)
library(viridis)
library(here)
library(tidyr)
library(scales)
library(patchwork)
```

# 1. Introduction

This document presents a data-driven approach for optimizing maintenance work ahead of a HUD NSPIRE (National Standards for the Physical Inspection of Real Estate) inspection. By utilizing mathematical techniques such as Monte Carlo sampling, the model runs thousands of "virtual inspections" to provide probabilistic estimations of key outcomes, such as the risk of regulatory failure.

To mirror real-world operations, the preparation work is modeled as a dependency-constrained network of tasks—a strict operational flowchart. This framework is highly adaptable and can be easily customized with configuration files to fit different building floorplans, tenant cleanliness levels, and inspection timelines.

Ultimately, the primary role of this modeling is prescriptive: it is designed to identify critical operational bottlenecks and the specific parameter thresholds that most heavily impact the probability of a successful inspection and overall preparation cost.



# 2. The NSPIRE Scoring Mechanism

The inspection divides a property into distinct locations and evaluates defects based on severity. 

## 2.1. Definitions and Sets
* **Severity:** $s \in \{\text{LT}, \text{S}, \text{STD}\}$ (Life-Threatening, Severe, Standard)
* **Location:** $l \in \{\text{Unit}, \text{Inside}, \text{Outside}\}$
* $N$: Total number of units in the property.
* $n(N)$: The HUD-mandated sample size function, where $n$ is a discrete function of $N$.
* $C_{s,l}$: True total count of defects existing in the property.
* $c_{s,l}$: Observed count of defects recorded during the inspection.
* $w_{s,l}$: Static point weight (penalty) assigned to a defect.

## 2.2. The Deduction Function
The total point deduction $D$ is a summation over all severities and locations. For Unit locations, the observed defects are normalized by the sample size $n(N)$, which effectively scales the deduction based on the inspection tier:

$$D = \sum_{s} \sum_{l \in \{\text{Inside}, \text{Outside}\}} (c_{s,l} \cdot w_{s,l}) + \sum_{s} \left( \frac{c_{s,\text{Unit}}}{n(N)} \cdot w_{s,\text{Unit}} \right)$$

## 2.3. The Raw Continuous Score
The base score originates at 100 and is bounded at zero using the Heaviside step function $H(x)$:

$$S_{raw} = (100 - D) \cdot H(100 - D)$$

## 2.4. The Operational Compliance Function (Failsafe)
A score $\ge 60$ mathematically passes, but the presence of *any* Life-Threatening (LT) defect forces an immediate 24-hour mitigation compliance failure. Let $n_{LT}$ represent the total observed LT defects:

$$n_{LT} = \sum_{l} c_{\text{LT},l}$$

$$S_{eff} = S_{raw} \cdot H(0.5 - n_{LT})$$

# 3. Probabilistic Sampling and Realizations

The inspection score is not a single, guaranteed number (a deterministic value). Instead, it is a **random variable**, representing the final outcome of many smaller random events—much like rolling a pair of dice. For example, whether the kitchen GFCI in Unit 1 happens to break today is a random event; whether the HUD inspector randomly chooses to inspect Unit 1 is another.

Because of this randomness, running a single simulation (rolling the dice once) is not very informative. However, by running thousands of virtual inspections, called Monte Carlo simulations, we can build a complete score distribution. This allows us to move beyond guessing and mathematically estimate precise probabilities, such as the exact risk of failing the inspection (scoring below 60).

## 3.1. Stochastic Unit Sampling
HUD inspection protocols dictate that units are randomly sampled. For a property with $N$ total units, the inspector chooses a subset $n(N)$ based on the tiered sampling schedule defined in Table 1.

| Total Units ($N$) | Inspected Units ($n$) |
| :--- | :--- |
| 1–5 | All units |
| 6–7 | 6 |
| 8 | 7 |
| 9–10 | 8 |
| 11–12 | 9 |
| 13–14 | 10 |
| 15–16 | 11 |
| 17–18 | 12 |
| 19–21 | 13 |
| 22–24 | 14 |
| 25–27 | 15 |
| 28–30 | 16 |
| 31–35 | 17 |
| 36–39 | 18 |
| 40–45 | 19 |
| 46–51 | 20 |
| 52–59 | 21 |
| 60–67 | 22 |
| 68–78 | 23 |
| 79–92 | 24 |
| 93–110 | 25 |
| 111+ | 32 |

In massive high-rise properties ($N > 100$), the inspection score tends to be highly stable because extreme outcomes naturally "average out" (Law of Large Numbers). However, for small to mid-sized properties ($N \leq 50$), this random sampling creates much larger relative variations. Because the inspector is pulling a random handful of units, the defects they happen to stumble upon ($c_{s,\text{Unit}}$) can be drastically better or worse than the actual overall condition of the building ($C_{s,\text{Unit}}$). This "luck of the draw" is exactly what drives the wide spread of possible final scores in small to mid-sized properties.

## 3.2. Realization of Deductions
Let $I_{u}$ be an indicator variable for unit $u$, that is, $I_u = 1$ if unit $u$ is selected for inspection, and $0$ otherwise. The observed defect count $c_{s,\text{Unit}}$ for a specific simulation realization $\omega$ is defined by the subset of selected units:

$$c_{s,\text{Unit}}(\omega) = \sum_{u=1}^{N} I_u(\omega) \cdot \mathbb{1}(\text{unit } u \text{ has defect } s)$$

The model calculates the realization $D(\omega)$ for each simulation $k$ in the Monte Carlo ensemble. This records the variance, i.e. all observed values, inherent in the sampling process, which is the primary driver of the "lottery effect" in mid-sized properties.

# 4. Resource Mapping and Graph Topology (DAG)

In property maintenance, repairs cannot always happen simultaneously or in random order. For example, you cannot paint a wall before patching the drywall, and you cannot patch the drywall before fixing the plumbing leak behind it.

To model this operational reality, we organize maintenance tasks into a Directed Acyclic Graph (DAG). In simpler terms, a DAG is a strict, one-way flowchart that maps out the required "order of operations." Each task (mathematically called a vertex, $V_i$) requires an expected execution time $t_i$ and a material cost $m_i$.

```{mermaid}
%%| fig-cap: "A simple Directed Acyclic Graph (DAG) illustrating mandatory task sequences."
flowchart LR
    A[Fix Plumbing Leak] --> B[Patch Drywall]
    B --> C[Wait for Compound to Dry]
    C --> D[Paint Wall]
    
    style A fill:#f9f9f9,stroke:#333,stroke-width:2px
    style B fill:#f9f9f9,stroke:#333,stroke-width:2px
    style C fill:#ffe6cc,stroke:#d79b00,stroke-width:2px,stroke-dasharray: 5 5
    style D fill:#f9f9f9,stroke:#333,stroke-width:2px
```

```{r, fig.width=10, fig.height=8}
#| echo: false
#| warning: false
#| message: false

library(igraph)
library(RColorBrewer)

# 1. Load the DAG
dag_csv <- "../inst/extdata/tasks_config.csv"
dag_obj <- create_inspection_dag(dag_csv)
g <- dag_obj@graph

# 2. Add Category-based colors
categories <- dag_obj@tasks_config$category
cat_levels <- unique(categories)
cols <- brewer.pal(max(3, length(cat_levels)), "Set3")
V(g)$color <- cols[match(categories, cat_levels)]

# 3. Size nodes by Betweenness
V(g)$size <- 8 + (betweenness(g) / max(betweenness(g)) * 15)

# 4. Refined Plotting
# par(mar) sets margins: c(bottom, left, top, right)
# Reducing top/bottom margins to expand plot and reduce title-to-plot gap
par(mar = c(1, 1, 2, 1)) 

plot(
  g,
  layout = layout_with_fr(g),
  vertex.label.cex = 1.5,
  vertex.label.color = "black",
  vertex.frame.color = "white",
  vertex.label.family = "sans",
  edge.width = 1,
  edge.arrow.size = 1,
  main = "", # We will add the title manually to control spacing better
  xlab = NA, ylab = NA
)

# Manually add title closer to the graph
title(main = "Maintenance Tasks DAG", line = 0.5, cex.main = 1.8)

# Legend at bottom left
legend("bottomleft", legend = cat_levels, col = cols, pch = 21, 
       pt.bg = cols, cex = 1.6, bty = "n")
```

## 4.1. Edge Constraints
1.  **Sequential Precedence:** A directed edge $V_A \to V_B$ requires the start time of $B$ to be $\ge$ the end time of $A$.
2.  **Temporal Blocking:** Certain vertices impose a local resource freeze, locking a spatial node for a time interval $\Delta t$.
3.  **Mutually Exclusive Pruning:** If a higher-order vertex $V_{replace}$ is activated, redundant child vertices $V_{repair}$ are pruned ($t_{repair} \to 0$).

# 5. Task Modalities & Execution Constraints

Task execution times are not normally distributed; they represent distinct physical realities (e.g., a quick adjustment versus a catastrophic replacement). 

## 5.1. Mixture Distributions (PERT)
Task $i$ exists in state $k \in \{1, 2, 3\}$. The execution time $t_i$ is drawn from a bounded Beta-PERT distribution, weighted by the physical probability of the state ($p_{i,k}$):

$$f(t_i) = \sum_{k} p_{i,k} \cdot \text{PERT}(t; a_k, m_k, b_k)$$
where $a_k, m_k, b_k$ represent the optimistic, mode, and pessimistic completion times.

## 5.2. Vendor Handoff Nodes
If a task evaluates to a severe state requiring external licensing or specialization (e.g., licensed plumbing or glazing), the node logic bifurcates:
* Internal labor time drops to zero ($t_{internal} \to 0$).
* A schedule lag is injected, drawing from a Uniform distribution: $L \sim \text{Uniform}(l_{min}, l_{max})$ days.
* A mandatory spatial lockout is triggered (e.g., 24-hour cure time for silicone), forcing the DAG to re-route internal technicians.

# 6. Quality Control & Redundancy

Execution of a task does not guarantee absolute resolution. A task has a baseline probability ($p_{err}$) of being missed or improperly repaired.

## 6.1. Audit Efficacy
We introduce a formal internal audit efficacy parameter, $\alpha_\text{audit} \in [0, 1]$, representing the probability that the pre-inspection internal team detects and successfully mitigates a defect. 

If a defect survives the initial repair attempt (with probability $p_{err}$), the probability that it remains undetected by the audit phase is:
$$P(\text{Survive}) = (1 - \alpha_\text{audit}) \cdot p_{err}$$

## 6.2. Technician Redundancy
* **Serial Execution (Single Technician):** The probability of a defect surviving into the inspection phase follows the logic in 6.1.
* **Parallel Execution (Dual Technicians):** Applying the "Four Eyes" principle, independent cross-checking drastically reduces error. The joint probability of both technicians missing the defect becomes $P(\text{Survive}) = (1 - \alpha_\text{audit}) \cdot p_{err}^2$.

# 7. Temporal Windowing and Asset Decay

The operational model assumes a maintenance repair window defined by $t_{start}$ (commencement of internal repairs) and $t_{end}$ (completion of repairs/audit). The interval $\delta = t_{start} - t_{end}$ constrains the total available labor capacity, while the subsequent interval $\delta_{decay} = t_{inspection} - t_{end}$ dictates the post-repair asset degradation.

## 7.1. Repair Capacity
Given a standard base labor capacity $R_{base}$ (hours/unit/day), the total labor capacity $\text{Cap}$ available for the maintenance cycle is linearly proportional to the duration of the repair window:


$$\text{Cap} = R_{base} \cdot (t_{start} - t_{end})$$

Tasks requiring labor hours $T_{total} > \text{Cap}$ result in a scheduling backlog, forcing the prioritization of high-penalty items (LT/S) at the expense of lower-severity items.

## 7.2. Tenant Heterogeneity (Bimodal Risk)
Asset deterioration is not uniform across a property. We model unit "cleanliness" as a bimodal latent state $\phi_u$ for each unit $u$, representing the inherent maintenance burden generated by the tenant. The tenant impact multiplier $\phi_u$ is drawn from a mixture distribution:

$$\phi_u \sim \begin{cases} \phi_{clean}, & \text{with probability } (1 - \pi) \\ \phi_{dirty}, & \text{with probability } \pi \end{cases}$$

Where $\pi$ is the estimated proportion of high-impact units in the property.

## 7.3. Post-Audit Asset Decay
Once repairs are audited at $t_{end}$, the asset re-enters a state of stochastic wear. We model the defect re-emergence as a Bernoulli process over the decay window $\delta_{decay}$. 

Given a daily probability of failure $\lambda_{i}$ for task $i$, the probability $P_{redefect}$ that a resolved task fails before the inspection date is:

$$P_{redefect} = 1 - (1 - \lambda_{i,u}^*)^{\delta_{decay}}$$

The effective daily probability of failure $\lambda_{i,u}^*$ is coupled with the unit's latent state $\phi_u$, effectively accelerating the re-emergence of defects in high-impact ("dirty") units:

$$\lambda_{i,u}^* = \lambda_{i} \cdot \phi_u$$

This temporal decay creates a trade-off: an earlier $t_{end}$ increases the decay window, inflating the probability that fixed items will fail before the inspector arrives, while a late $t_{end}$ constricts the repair window, potentially inducing a labor backlog.

# 8. Predictive Topological Risk (Latent Covariates)

Baseline task probabilities assume a standard asset deterioration rate. However, the lack of historical tenant reporting often masks hidden deferred maintenance, warping the DAG's risk topology. We use historical service request volume $W_u$ as a proxy to estimate the probability that a unit belongs to the "dirty" latent state ($\phi_{dirty}$).

## 8.1. Inferring Latent States
We define the unit-specific probability $\pi_u$ of a unit being in the high-impact state as a logistic function of service requests $W_u$:

$$\pi_u = P(\phi_u = \phi_{dirty} | W_u) = \text{logit}^{-1}(\alpha - \beta W_u)$$

Where $\alpha$ and $\beta$ are calibrated tuning parameters. For units with low reporting ($W_u \to 0$), the prior probability of being in the "dirty" state increases, signaling a higher likelihood of latent deferred maintenance.

## 8.2. Dynamic Risk Mapping

This unit-specific $\pi_u$ dynamically replaces the global mixture probability used in Section 7.2. During simulation initialization, each unit's tenant factor $\phi_u$ is sampled conditioned on its specific reporting history:

$$\phi_u \sim \begin{cases} \phi_{clean}, & \text{with probability } (1 - \pi_u) \\ \phi_{dirty}, & \text{with probability } \pi_u \end{cases}$$

This approach transforms the simulation from a static property model into a predictive prioritization tool. By mapping historical service request volumes to the bimodal risk topology, the engine directs the maintenance DAG to prioritize units where $W_u$ is low (indicating a high probability of "hidden" defects) before they enter the randomized HUD sampling window.

```{r}
#| label: tbl-tasks-config
#| tbl-cap: "NSPIRE Task Modalities and Execution Constraints"
#| echo: false
#| warning: false
#| message: false

# Load the required libraries
library(DT)
library(readr)

# Read the CSV
tasks_df <- read_csv("../inst/extdata/tasks_config.csv")

# Generate and print the interactive table
datatable(
  tasks_df,
  extensions = c('FixedColumns', 'Responsive'),
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    fixedColumns = list(leftColumns = 5), 
    autoWidth = TRUE,
    columnDefs = list(
      list(className = 'dt-center', targets = "_all")
    )
  ),
  class = 'cell-border stripe hover'
)
```

# 9. Risk Analysis

Moving beyond deterministic score projections, this model quantifies operational risk by treating property performance as a stochastic process. By leveraging Monte Carlo simulation, we assess not only the expected outcome but the probability of regulatory non-compliance.

## 9.1. Quantification of Failure Risk

We define the failure event as $S_{eff} < 60$. Due to the non-linear interaction of random unit sampling, defect emergence probabilities, and repair quality, the inspection score $S_{eff}$ is a random variable. We estimate the probability of failure ($P_{fail}$) by simulating $K = 1,000$ independent inspection realizations:

$$P_{fail} = \frac{1}{K} \sum_{i=1}^{K} I(S_{eff,i} < 60)$$

where $I(\cdot)$ is the indicator function. This allows us to map the property's risk profile against the HUD enforcement threshold, providing a granular view of the likelihood of catastrophic failure that a mean-based score effectively masks.

## 9.2. Quantifying Volatility: The Tail Drop Metric

Standard deviation ($\sigma$) and the Coefficient of Variation ($CV = \sigma/\mu$) provide a global measure of dispersion. However, in NSPIRE compliance, the right tail (scores $> 100$) is truncated, and the distribution is often skewed. The primary operational risk resides in the left tail—the possibility of a "bad day" where sampling captures a cluster of high-severity defects.

To quantify this downside volatility, we define the **Tail Drop Metric** ($\tau$):

$$\tau = \frac{\text{med} - P_{05}}{\text{med}}$$ {#eq-tau}

Where:

*   $\text{med}$ is the median score.
*   $P_{05}$ is the 5th percentile of the simulated scores (the "Value at Risk").

$\tau$ represents the percentage degradation from expected performance to the worst-case scenario. A property may exhibit a high $\text{med}$ while maintaining a high $\tau$, identifying it as a "lottery" asset—stable on average, but highly susceptible to non-linear drops in score due to the interaction of unit sampling variance and tenant-driven decay.

## 9.3. Sensitivity Mapping

The Tail Drop Metric $\tau$ is a systemic output of our model, functionally defined by the interaction of our control and stochastic parameters:
$$\tau = f(n(N), \alpha_\text{audit}, \pi, \phi_{dirty})$$

This sensitivity function captures how variance in the HUD sampling tier $n(N)$, human error in pre-inspection audit ($\alpha_\text{audit}$), and the underlying distribution of tenant impact ($\pi, \phi_{dirty}$) propagate into downside performance risk. Future numerical investigations will isolate these partial derivatives to determine which parameter dominates the variance in the $P_{05}$ left tail.

---

# 10. Parameter Sensitivity Analysis

Sensitivity analysis studies how variations in control parameters drive system outcomes, such as expected score, labor costs, and downside risk ($P_{fail}$ and $\tau$). By studying these variables across parameter ranges, one can identify parameter critical ranges, non-linear thresholds and propose optimal resource allocations within the NSPIRE framework.

## 10.1. Audit Catch Rate ($\alpha_\text{audit}$)

The internal pre-inspection audit is modeled by $\alpha_\text{audit}$, the probability that a defect is identified and resolved prior to the HUD inspection. To quantify the impact of audit efficacy on compliance risk, $\alpha_\text{audit}$ was varied from $0.05$ (negligent) to $0.95$ (highly rigorous), while executing $K=1,000$ Monte Carlo iterations per step. @fig-audit-ridges provides a visual summary of the resulting score distributions (horizontal axis with a red line for the failing score of 60) as a function of $\alpha_\text{audit}$ (vertical axis).

```{r}
#| label: fig-audit-ridges
#| fig-width: 9
#| fig-height: 6
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Risk Transition: Audit Rate vs. Score Distribution."

sweep_data <- load_sensitivity_iterations(here("results", "sensitivity", "audit_rate"))

# Generate the plot
plot_sensitivity_ridges(
  sensitivity_df = sweep_data,
  param_label = "Audit Catch Rate (Efficacy)",
  title = "Audit Rate vs. Score Distribution"
)
```

Failure risk is almost completely eliminated when $\alpha_\text{audit} \geq 0.9$. @fig-distribution-slice further explains the transition from high risk to low risk as $\alpha_\text{audit}$ increases from 0.55 to 0.85. The failure risk is best estimated with the proportion of simulations (among $K=1,000$) that yield a score below 60 (red vertical line). Visually, this corresponds to the area under the density curve (integral) to the left of the critical red line. 


```{r} 
#| label: fig-distribution-slice
#| fig.width: 12 
#| fig.height: 6
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Examples of Score Distribution: $\\alpha_\\text{audit} = 0.55$ (left) and $\\alpha_\\text{audit} = 0.85$ (right)."

# 1. Define the scenarios
scenarios <- c(0.55, 0.85)
sum_df <- readRDS(here("results", "sensitivity", "audit_rate", "summary.rds"))
pal <- scales::viridis_pal(option = "D")(100) # Pre-generate palette

plots <- lapply(scenarios, function(rate) {
  # 1. Cast both to numeric to ensure they match
  median_val <- sum_df$score_median[as.numeric(sum_df$param_val) == as.numeric(rate)]
  
  # 2. Pick color based on median score (0-100 mapped to 1-100 index)
  idx <- max(1, min(100, round(median_val)))
  
  plot_score_distribution(
    input = here("results", "sensitivity", "audit_rate", paste0("score_density_", rate, ".rds")),
    title = paste("Audit Catch Rate:", rate * 100, "%"),
    color_hex = pal[idx]
  )
})

plots[[1]] | plots[[2]]

```

Estimating both failure risks (score less than 30 or less than 60) at all values of $\alpha_\text{audit}$ yields @fig-failure-audit. The data demonstrates that critical failure risk (score < 30) is only marginalized when audit catch rates exceed 0.75, while full protection against standard failure (score < 60) requires efficacy above 0.90.

```{r}
#| label: fig-failure-audit
#| fig.width: 6
#| fig.height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Failure probabilities at scores 30 and 60 as a function of audit catch rate $\\alpha_\\text{audit}$."

sum_df <- readRDS(here("results", "sensitivity", "audit_rate", "summary.rds"))

plot_sensitivity_multi(
  sum_df, 
  y_vars = c("fail_risk_60", "critical_fail_risk_30"),
  param_label = "Audit Catch Rate",
  y_label = "Probability",
  title = "Impact of Audit Efficacy on Regulatory Failure"
)
```

Besides failure probabilities, another useful statistic is the tail-drop metric $\tau$ @eq-tau. To understand its sensitivity to $\alpha_\text{audit}$, we must first examine how the foundational score statistics evolve across the parameter range. These plots are displayed in @fig-dist-stats-audit.

```{r}
#| label: fig-dist-stats-audit
#| fig.width: 6
#| fig.height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Score statistics as a function of audit catch rate $\\alpha_\\text{audit}$."

sum_df <- readRDS(here("results", "sensitivity", "audit_rate", "summary.rds"))

plot_sensitivity_multi(
  sum_df, 
  y_vars = c("score_p05", "score_median", "score_sd"),
  param_label = "Audit Catch Rate",
  y_label = "Score",
  title = "Impact of Audit Efficacy on Score Statistics"
)
```

As $\alpha_\text{audit}$ increases beyond 0.25, the score median starts rising linearly. Later, as $\alpha_\text{audit}$ gets above 0.45, the 5th percentile score $P_{05}$ also rises linearly but with a stiffer slope. This yields decreasing differences between median and 5th percentile. One way to visualize this is to plot the standard deviation of the score (a proxy measure for difference between median and 5th percentile) as a function of $\alpha_\text{audit}$. These behaviors of the different statistics yield the profile of $\tau$ @eq-tau as a function of $\alpha_\text{audit}$ displayed in @fig-tau-audit.

```{r}
#| label: fig-tau-audit
#| fig.width: 6
#| fig.height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Tail drop metric $\\tau$ as a function of audit catch rate $\\alpha_\\text{audit}$."

sum_df <- readRDS(here("results", "sensitivity", "audit_rate", "summary.rds"))

plot_sensitivity_multi(
  sum_df, 
  y_vars = c("tail_drop_median"),
  param_label = "Audit Catch Rate",
  y_label = "Tail Drop Metric",
  title = "Impact of Audit Efficacy on Tail Drop Metric."
)
```

**Conclusions for $\alpha_\text{audit}$**:

* Operational Futility Threshold ($\alpha_\text{audit} \leq 0.25$): Below ~25%
efficacy, quality assurance efforts are mathematically invisible to the final metric due to the absolute scoring floor (0).

* Variance Squeeze ($0.45 < \alpha_\text{audit} < 0.90$): As the catch rate rises above 0.45, the 5th percentile worst-case outcome ($P_{05}$) improves at a faster linear rate than the median. This compresses the score variance and significantly reduces the Tail Drop metric ($\tau$).

* Safety Baseline ($\alpha_\text{audit} \geq 0.90$): To confidently eliminate standard regulatory failure risk (score < 60), the internal audit process must successfully intercept at least 90% of all pre-existing defects.

---

## 10.2. The Time Crunch: Notice of Inspection ($t_\text{start}$)

While a robust audit catch rate ($\alpha_\text{audit}$) is critical for passing an NSPIRE inspection, it assumes the maintenance team has the physical time to execute the repairs. HUD inspection notices can vary drastically, sometimes providing as little as 14 to 30 days of warning.

To model this operational squeeze, the preparation window ($t_\text{start}$) was swept from 36 days down to 4 days. For this specific scenario, the internal audit catch rate was fixed at a highly effective 85% ($\alpha_\text{audit} = 0.85$), and maximum allowable overtime was expanded to 4 hours per unit per day. Crucially, the simulation incorporates context-aware transit friction: technicians incur small delays (1–5 minutes) when moving between tasks in the same unit, but significant delays (15–30 minutes) when packing up and accessing a new unit.

As the preparation window shrinks, the volume of identified defects remains constant, but the time available to clear them—and the time available to walk between them—compresses. 

```{r}
#| label: fig-time-crunch-ridges
#| fig-width: 9
#| fig-height: 6
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Score Distribution Evolution under time compression. Ridges are shown at 4-day intervals for visual clarity."

sweep_data <- load_sensitivity_iterations(here("results", "sensitivity", "t_start"))

# Subset to avoid visual clutter (e.g., plotting every 4 days)
subset_days <- seq(4, 36, by = 3)
sweep_data_sub <- sweep_data[as.numeric(sweep_data$param_val) %in% subset_days, ]

# Generate the plot
plot_sensitivity_ridges(
  sensitivity_df = sweep_data_sub,
  param_label = "Days of Notice (t_start)",
  title = "Time Crunch vs. Score Distribution"
)
```

As visualised in @fig-time-crunch-ridges, the score distribution remains stable when $t_\text{start} \geq 28$. However, as time compresses, the distribution stretches heavily to the left before completely fracturing and collapsing into the failure zone.

The initial impact is purely financial, as visualized in @fig-time-crunch-labor.

```{r}
#| label: fig-time-crunch-labor
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Mean total labor cost and overtime cost as the preparation window ($t_\\text{start}$) changes. Note the massive overtime spike at Day 22 followed by the sharp drop in total costs as the timeline collapses."

sum_df_time <- readRDS(here("results", "sensitivity", "t_start", "master_time_crunch_study.rds"))

plot_sensitivity_multi(
  sum_df_time, 
  y_vars = c("total_cost_mean", "overtime_mean"),
  param_label = "Days of Notice (t_start)",
  y_label = "Cost ($)",
  title = "Financial Impact of Compressed Prep Windows"
)
```

At first glance, the left side of @fig-time-crunch-labor appears counter-intuitive: as the preparation window drops below 20 days, the total expected cost plunges dramatically. This is not an efficiency gain; it is the mathematical manifestation of deferred maintenance.

Between Day 30 and Day 22, the workforce heavily leverages their overtime allowance to keep pace with the workload. At Day 22, the system hits peak labor output—the team is operating at absolute maximum capacity, driving overtime spend to its highest point. However, as the window shrinks below 20 days, the absolute physical capacity of the workforce is exhausted. Hundreds of items are forced into the "backlog." Because backlogged items are abandoned, they incur no labor and no material costs, creating a "false economy" of savings.

@fig-time-crunch-backlog illustrates the severe regulatory consequences of this capacity exhaustion.

```{r}
#| label: fig-time-crunch-backlog
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Probability of generating a task backlog and the resulting regulatory failure risk ($<60$) under compressed timelines."

plot_sensitivity_multi(
  sum_df_time, 
  y_vars = c("backlog_prob", "fail_risk_60", "critical_fail_risk_30"),
  param_label = "Days of Notice (t_start)",
  y_label = "Probability",
  title = "The Backlog Cliff: Capacity Exhaustion"
)
```

Finally, we map the Tail Drop metric ($\tau$) as a function of the preparation window. This demonstrates how temporal limits affect the volatility of outcomes.

```{r}
#| label: fig-time-crunch-tau
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Tail Drop Metric $\\tau$ as a function of the preparation window. High $\\tau$ indicates massive outcome volatility."

plot_sensitivity_multi(
  sum_df_time, 
  y_vars = c("tail_drop_median"),
  param_label = "Days of Notice (t_start)",
  y_label = "Tail Drop Metric",
  title = "Impact of Prep Window on Tail Drop Metric"
)
```

**Conclusions for $t_\text{start}$**:

**Distributed Defect Tax ($t_\text{start} \geq 28$)**: Because repairing widely distributed defects incurs a heavy transit-time penalty, a 28-day notice is the edge of the safe operational limit for this property. At 28 days, the property maintains a near 0% failure risk with minimal overtime expenditure.

**Maximum Effort Peak ($t_\text{start} \approx 22$)**: Day 22 represents the system's breaking point. Overtime spend reaches its absolute maximum as the team attempts to outwork the shrinking timeline. Despite spending heavily on premium labor, the backlog probability reaches ~50%, and the property begins to exhibit a measurable risk of regulatory failure. The team is throwing maximum capital at the problem but is still losing ground.

**Operational Collapse ($t_\text{start} \leq 15$)**: A standard 14-day HUD notice is mathematically unfeasible for a property in this state and with 5 technicians. By Day 15, the probability of backlogging tasks reaches 100%, and the property's risk of failure exceeds 85%. The simulation proves that sudden drops in preparation expenditure in the days leading up to an inspection are a lagging indicator of capacity exhaustion, representing a forced trade-off of immediate OpEx savings for guaranteed regulatory failure.


## 10.3. Prescriptive Analytics: The Staffing Curve

Having established in Section 10.2 that a compressed timeline guarantees regulatory failure due to capacity exhaustion, we can invert the simulation to perform prescriptive analytics. By decoupling the labor constraints from the QA blind spot, we can query the engine: Given a specific preparation window, what is the exact minimum headcount required to guarantee a 99% probability of clearing the known backlog?

Using a binary search algorithm wrapped around the Monte Carlo engine (with $K=1,000$ iterations per step), we mapped the required workforce capacity as the timeline collapses. We targeted a strict operational threshold ($P(\text{Backlog} > 0) \le 0.01$) to ensure the recommended staffing levels are mathematically robust.

```{r}
#| label: fig-staffing-curve
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Prescriptive Staffing Curve: Minimum technicians required to guarantee $P(\\text{Backlog}) \\le 0.01$ as the inspection window compresses."

# Load the high-resolution staffing curve data
# Assuming it was saved via: saveRDS(staffing_df, here("results", "sensitivity", "t_start", "staffing_curve.rds"))
staffing_df <- readRDS(here::here("results", "sensitivity", "t_start", "staffing_curve.rds"))

ggplot2::ggplot(staffing_df, ggplot2::aes(x = t_start, y = required_techs)) +
  ggplot2::geom_line(color = "#8b0000", linewidth = 1.2, direction = "vh") +
  ggplot2::geom_point(shape = 21, fill = "white", color = "#8b0000", size = 2, stroke = 1.2) +
  ggplot2::scale_x_reverse() + # Reverse X-axis so time counts down to the right
  ggplot2::scale_y_continuous(breaks = scales::pretty_breaks(n = 6)) +
  ggplot2::theme_minimal(base_size = 14) +
  ggplot2::labs(
    title = "Prescriptive Staffing Curve",
    x = "Days of Notice (t_start)",
    y = "Required Technicians"
  ) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    plot.title = ggplot2::element_text(face = "bold")
  )
```

The resulting curve is highly non-linear. With a generous notice period, the baseline crew of technicians is perfectly adequate to absorb the transit friction and complete the repairs. However, dropping below the 15-day threshold triggers an exponential surge in labor demand. To prep this property in under a week, a property manager cannot simply authorize overtime; they must instantly deploy a massive, coordinated swarm of technicians.

This prescriptive tool translates abstract regulatory timelines directly into actionable HR and budgeting directives, allowing management to definitively calculate the cost of a delayed response.

---

## 10.4. Post-Prep Decay: The "Too Early" Penalty ($t_\text{end}$)

In Section 10.2, we established that compressing the preparation timeline forces a property into a backlog cliff due to labor limits. A common operational response to this risk is to overcompensate by completing all pre-inspection repairs well in advance.

However, residential properties are dynamic, entropic environments subject to continuous wear-and-tear. The time gap between the completion of repairs ($t_\text{end}$) and the actual day of the HUD inspection exposes the property to a natural re-accumulation of defects.

To quantify this temporal decay risk, we swept $t_\text{end}$ from 0 days (repairs finish the morning of the inspection) up to 35 days. To isolate the effect of decay from capacity constraints, the starting notice ($t_\text{start}$) was locked at 45 days, ensuring the team had abundant capacity and zero backlog risk.

@fig-decay-score illustrates the erosion of the property's safety margin as the time gap widens.

```{r}
#| label: fig-decay-score
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Erosion of the Median and 5th Percentile ($P_{05}$) scores as the gap between prep completion and the inspection date widens."

sum_df_decay <- readRDS(here::here("results", "sensitivity", "t_end", "master_decay_study.rds"))

plot_sensitivity_multi(
  sum_df_decay, 
  y_vars = c("score_median", "score_p05"),
  param_label = "Days Finished Before Inspection (t_end)",
  y_label = "Score",
  title = "Impact of Post-Prep Decay on Score Confidence"
)
```

As visualized in @fig-decay-score, the score does not degrade at a steady, linear rate. Instead, the property exhibits strong resilience for the first three weeks, maintaining stable median and $P_{05}$ scores up to $t_\text{end} \approx 25$. However, beyond 25 days, the score plunges rapidly. This indicates that temporal decay acts as a threshold effect: accumulating minor wear-and-tear is initially absorbable, but as time passes, the compounding probability of a critical life-safety system (e.g., a smoke detector or GFCI outlet) failing at random reaches a critical mass, shattering the score.

@fig-decay-risk visualizes how this sudden score erosion translates into concrete regulatory failure.

```{r}
#| label: fig-decay-risk
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "The compounding probability of regulatory failure caused by natural property decay during the post-prep waiting period."

plot_sensitivity_multi(
  sum_df_decay, 
  y_vars = c("fail_risk_60", "critical_fail_risk_30"),
  param_label = "Days Finished Before Inspection (t_end)",
  y_label = "Probability",
  title = "Failure Risk Amplification over Time"
)
```

**Conclusions for $t_\text{end}$**:

**Decay Cliff**: Completing preparations too early introduces severe mathematical risk, manifesting as a "cliff" rather than a steady slope. The data demonstrates that a property remains relatively safe if it sits idle for up to three weeks. However, a property that is highly likely to pass at $t_\text{end} = 0$ exhibits an exponential spike in failure risk if left idle beyond 25 days, reaching near-certain failure by day 35.

**"Goldilocks" Prep Window**: Combined with the capacity insights from Section 10.2, these results define a rigid operational window. A maintenance team must start late enough to avoid an excessive decay penalty ($t_\text{end} \le 25$ days) but early enough to avoid capacity exhaustion ($t_\text{start}$ constraint). Maximizing NSPIRE performance requires precise "just-in-time" repair scheduling.

---

## 10.5. Spatial Heterogeneity and the "Lottery Effect"

NSPIRE scores are heavily influenced by the spatial distribution of defects. HUD's sampling protocol selects a subset of units to represent the entire property (e.g., sampling 20 units in a 50-unit building). While this works mathematically for uniformly maintained properties, it introduces massive variance—the "Lottery Effect"—in properties with severe tenant heterogeneity.

To quantify this, we swept the clean_proportion parameter from 0.50 (highly heterogeneous: 50% normal tenants, 50% highly destructive tenants) up to 1.00 (homogeneous: 100% normal tenants). To isolate the sampling risk, the "destructive" tenants were assigned a severe 5.0x wear-and-tear multiplier, and the internal audit catch rate was fixed at 75%.

@fig-tenant-score visualizes how the introduction of destructive tenants fractures the property's safety margin.

```{r}
#| label: fig-tenant-score
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Median and 5th Percentile ($P_{05}$) scores as the proportion of clean/normal tenants increases. A highly heterogeneous building (left) exhibits massive tail drop."

sum_df_tenant <- readRDS(here::here("results", "sensitivity", "tenant_mix", "master_tenant_study.rds"))

plot_sensitivity_multi(
  sum_df_tenant, 
  y_vars = c("score_median", "score_p05"),
  param_label = "Proportion of Clean/Normal Tenants",
  y_label = "Score",
  title = "Impact of Tenant Heterogeneity on Score Confidence"
)
```

As the proportion of clean tenants drops below 85%, the 5th percentile worst-case scenario ($P_{05}$) decouples completely from the median, rapidly plunging toward zero. This decoupling is the direct result of sampling variance, which is best visualized by plotting the standard deviation of the inspection scores (@fig-tenant-variance).

```{r}
#| label: fig-tenant-variance
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "The 'Lottery Effect': Score volatility (Standard Deviation) explodes as the building becomes more heterogeneous."

plot_sensitivity_multi(
  sum_df_tenant, 
  y_vars = c("score_sd"),
  param_label = "Proportion of Clean/Normal Tenants",
  y_label = "Score Standard Deviation",
  title = "Score Volatility vs. Tenant Heterogeneity"
)
```

@fig-tenant-variance provides mathematical proof of the Lottery Effect. In a homogeneous building (clean_proportion = 1.0), the score is highly predictable (Standard Deviation $\approx 7$). However, in a heterogeneous building (clean_proportion = 0.60), the standard deviation triples to over 21 points.

In this state, the final NSPIRE score is dictated entirely by random chance. If the HUD inspector's random sample happens to bypass the highly destructive units, the property may easily pass. If the sample includes several of them, the heavy penalties (e.g., 60.0 pre-division points for Life-Threatening items) instantly drag the score below 60.

@fig-tenant-risk demonstrates how this volatility translates into regulatory failure.

```{r}
#| label: fig-tenant-risk
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "The probability of regulatory failure as a function of tenant composition."

plot_sensitivity_multi(
  sum_df_tenant, 
  y_vars = c("fail_risk_60", "critical_fail_risk_30"),
  param_label = "Proportion of Clean/Normal Tenants",
  y_label = "Probability",
  title = "Failure Risk Driven by High-Impact Tenants"
)
```

**Conclusions for Tenant Heterogeneity**:

**Lottery Effect**: Under the NSPIRE scoring protocol, you cannot "average out" a destructive tenant. Because severe defects carry extreme penalty weights, a small cluster of highly degraded units acts as a regulatory landmine.

**Targeted Pre-Inspections**: This proves that treating all units equally during the preparation window is sub-optimal. If management is aware of specific high-impact tenants, those specific units must be audited and repaired at a 100% catch rate prior to the inspection, as they pose a disproportionate mathematical threat to the entire property's score.

---

<style>
.floating-copyright {
  position: fixed;
  bottom: 15px;
  left: 15px;
  font-size: 0.85em;
  color: #777;
  background-color: rgba(255, 255, 255, 0.85);
  padding: 5px 10px;
  border-radius: 5px;
  z-index: 1000;
  pointer-events: none; 
}
</style>

<div class="floating-copyright">
  &copy; 2026 Joël R. Pradines.<br> All rights reserved.
</div>