---
title: "Optimization of Pre-NSPIRE Inspection Work via Monte Carlo Simulation"
author: "Joël R. Pradines, Ph.D."
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
    math: true
    embed-resources: true
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

devtools::load_all(".")
library(ggplot2)
library(ggridges)
library(viridis)
library(here)
library(tidyr)
library(scales)
library(patchwork)
```

# 1. Introduction

Preparing a property for a HUD NSPIRE inspection is a complex operational challenge governed by rigid constraints: labor limits, task dependencies, and tenant behavior.

This document presents a data-driven approach to optimizing this pre-inspection workflow. By leveraging Monte Carlo simulation, we execute thousands of "virtual inspections." This stochastic engine stress-tests a property's risk profile, bridging ground-level maintenance realities with probabilistic modeling. To enforce the physical laws of maintenance, the workflow is mapped as a Directed Acyclic Graph (DAG), ensuring tasks sequence correctly and temporal lockouts are respected.

Crucially, this model is not a crystal ball. It is not designed to predict a deterministic score for a specific building. Rather, its immediate value is prescriptive. It isolates the mathematical architecture of an inspection to identify critical operational levers and their specific failure thresholds.

Ultimately, this document serves as a proof-of-concept for the utility of quantitative analysis in property management. By mapping the interaction of time, labor, and probability, it demonstrates how raw operational data can be translated into definitive directives—identifying staffing requirements, exposing the limits of overtime, and defining the precise timelines required to guarantee regulatory compliance. Specifically, the simulation reveals the following key findings:

* **Hidden Labor Tax**: Incorporating realistic human behavior—such as 5-minute intra-unit tool transitions and 15-minute inter-unit transit times—imposes a rigid 33% labor tax on baseline time estimates.
* **Quality Assurance Floor**: Internal pre-inspections must intercept at least 90% of defects to eliminate failure risk. Conversely, an audit catch rate below 25% is mathematically futile, offering zero protection against the scoring floor.
* **Overtime Illusion**: Compressing preparation timelines below 15 days forces the workforce into a "backlog cliff." At this threshold, maximum capital expenditure on overtime cannot prevent regulatory failure.
* **Non-Linear Staffing Surges**: Shortened inspection notices cannot be mitigated simply by authorizing more hours; instead, they demand an exponential surge in active headcount.
* **Post-Prep Decay Penalty**: Completing repairs too early introduces severe physical decay risk. Properties sitting idle for more than 25 days before an inspection face an exponential spike in defect re-emergence.
* **Lottery Effect**: HUD’s random sampling protocol severely penalizes spatial variance. A handful of highly destructive tenants in an otherwise pristine building triples the probability of a catastrophic inspection failure. To neutralize this threat, management must aggressively target high-risk units during the internal audit, using low historical service request rates as a predictive proxy for hidden defects.

# 2. The NSPIRE Scoring Mechanism

The inspection divides a property into distinct locations and evaluates defects based on severity. 

## 2.1. Definitions and Sets

* **Severity:** $s \in \{\text{LT}, \text{S}, \text{STD}\}$ (Life-Threatening, Severe, Standard)
* **Location:** $l \in \{\text{Unit}, \text{Inside}, \text{Outside}\}$
* $N$: Total number of units in the property.
* $n(N)$: The HUD-mandated sample size function, where $n$ is a discrete function of $N$.
* $C_{s,l}$: True total count of defects existing in the property.
* $c_{s,l}$: Observed count of defects recorded during the inspection.
* $w_{s,l}$: Static point weight (penalty) assigned to a defect.

## 2.2. Deduction Function

The total point deduction $D$ is a summation over all severities and locations. For Unit locations, the observed defects are normalized by the sample size $n(N)$, which effectively scales the deduction based on the inspection tier:

$$D = \sum_{s} \sum_{l \in \{\text{Inside}, \text{Outside}\}} (c_{s,l} \cdot w_{s,l}) + \sum_{s} \left( \frac{c_{s,\text{Unit}}}{n(N)} \cdot w_{s,\text{Unit}} \right)$$

## 2.3. Raw Continuous Score

The base score originates at 100 and is bounded at zero using the Heaviside step function $H(x)$:

$$S_{raw} = (100 - D) \cdot H(100 - D)$$

## 2.4. Operational Compliance Function (Failsafe)

A score $\ge 60$ mathematically passes, but the presence of **any** Life-Threatening (LT) defect forces an immediate 24-hour mitigation compliance failure. Let $n_{LT}$ represent the total observed LT defects:

$$n_{LT} = \sum_{l} c_{\text{LT},l}$$.

We model this regulatory mandate as a mathematical "kill switch" called the Heaviside function $H(x)$, defined by:
$$H(x) = \begin{cases} 1 & \text{if } x \ge 0 \\ 0 & \text{otherwise} \end{cases}$$
If even one LT defect is observed ($n_{LT} \ge 1$), the term $(0.5 - n_{LT})$ becomes negative. This drives the function to output a zero, instantly failing the property regardless of its raw score:
$$S_{eff} = S_{raw} \cdot H(0.5 - n_{LT})$$

# 3. Probabilistic Sampling and Realizations

The inspection score is not a single, guaranteed number (a deterministic value). Instead, it is a **random variable**, representing the final outcome of many smaller random events—much like rolling a pair of dice. For example, whether the kitchen GFCI in Unit 1 happens to break today is a random event; whether the HUD inspector randomly chooses to inspect Unit 1 is another.

Because of this randomness, running a single simulation (rolling the dice once) is not very informative. However, by running thousands of virtual inspections, called Monte Carlo simulations, we can build a complete score distribution. This allows us to move beyond guessing and mathematically estimate precise probabilities, such as the risk of failing the inspection (scoring below 60).

## 3.1. Stochastic Unit Sampling

HUD inspection protocols dictate that units are randomly sampled. For a property with $N$ total units, the inspector chooses a subset $n(N)$ based on the tiered sampling schedule defined in Table 1.

```{r}
#| label: tbl-hud-sampling
#| tbl-cap: "HUD NSPIRE Tiered Sampling Schedule"
#| echo: false
#| warning: false
#| message: false

library(DT)

# 1. Base Data
hud_sampling <- data.frame(
  `Total Units (N)` = c("1–5", "6–7", "8", "9–10", "11–12", "13–14", "15–16", "17–18", 
                        "19–21", "22–24", "25–27", "28–30", "31–35", "36–39", "40–45", 
                        "46–51", "52–59", "60–67", "68–78", "79–92", "93–110", "111+"),
  `Inspected Units (n)` = c("All units", "6", "7", "8", "9", "10", "11", "12", "13", 
                            "14", "15", "16", "17", "18", "19", "20", "21", "22", "23", 
                            "24", "25", "32"),
  check.names = FALSE
)

# 2. Define mathematical boundaries
min_n <- c(1, 6, 8, 9, 11, 13, 15, 17, 19, 22, 25, 28, 31, 36, 40, 46, 52, 60, 68, 79, 93, 111)
max_n <- c(5, 7, 8, 10, 12, 14, 16, 18, 21, 24, 27, 30, 35, 39, 45, 51, 59, 67, 78, 92, 110, 500)

# 3. Create the hidden search index string (e.g., "25 26 27")
hud_sampling$SearchTerms <- mapply(function(start, end) {
  paste(seq(start, end), collapse = " ")
}, min_n, max_n)

# 4. Render Interactive Table with custom Regex search
datatable(
  hud_sampling,
  options = list(
    pageLength = 8,
    dom = 'ftp', 
    autoWidth = TRUE,
    columnDefs = list(
      list(className = 'dt-center', targets = 0:1),
      list(visible = FALSE, targets = 2) # Hide the search string
    ),
    # Inject JavaScript to wrap the user's input in word boundaries (\b)
    initComplete = JS(
      "function(settings, json) {",
      "  var dt = this.api();",
      "  $('.dataTables_filter input').unbind().on('keyup', function(e) {",
      "    var val = this.value.trim();",
      "    if(val) {",
      "      dt.search('\\\\b' + val + '\\\\b', true, false).draw();",
      "    } else {",
      "      dt.search('').draw();",
      "    }",
      "  });",
      "}"
    )
  ),
  rownames = FALSE,
  class = 'cell-border stripe hover'
)
```

In massive high-rise properties ($N > 100$), the inspection score tends to be highly stable because extreme outcomes naturally "average out" (Law of Large Numbers). However, for small to mid-sized properties ($N \leq 50$), this random sampling creates much larger relative variations. Because the inspector is pulling a random handful of units, the defects they happen to stumble upon ($c_{s,\text{Unit}}$) can be drastically better or worse than the actual overall condition of the building ($C_{s,\text{Unit}}$). This "luck of the draw" is exactly what drives the wide spread of possible final scores in small to mid-sized properties.

## 3.2. Realization of Deductions

Let $I_{u}$ be an indicator variable for unit $u$, that is, $I_u = 1$ if unit $u$ is selected for inspection, and $0$ otherwise. The observed defect count $c_{s,\text{Unit}}$ for a specific simulation realization $\omega$ is defined by the subset of selected units:

$$c_{s,\text{Unit}}(\omega) = \sum_{u=1}^{N} I_u(\omega) \cdot \mathbb{1}(\text{unit } u \text{ has defect } s)$$

The model calculates the realization $D(\omega)$ for each simulation $k$ in the Monte Carlo ensemble. This records the variance, i.e. all observed values, inherent in the sampling process, which is the primary driver of the "lottery effect" in mid-sized properties.

## 3.3. Task Execution Times and Mixture Distributions

Just as the inspector's choice of units is random, the time required to complete a maintenance task is never a fixed number.

To model this physical reality, we treat execution time $t_i$ as a random variable drawn from a bounded Beta-PERT (Program Evaluation and Review Technique) distribution. PERT is widely used in project management because it relies on three intuitive estimates: an optimistic minimum ($a$), a most likely mode ($m$), and a pessimistic maximum ($b$).

However, tasks often exist in distinctly different physical states. Consider a sink drain: it might simply require a quick manual adjustment by a technician (State 1), or it might require a licensed plumber to replace the downstream PVC P-trap (State 2). To account for this, the execution time $t_i$ is drawn from a mixture distribution, weighted by the physical probability of each state ($p_{i,k}$):
$$f(t_i) = \sum_{k} p_{i,k} \cdot \text{PERT}(t; a_k, m_k, b_k)$$
Using a single "average time" for such a task would be an operational disaster, as it would plan for an event that never actually happens. As illustrated in @fig-multimodal-pert, the mixture distribution correctly captures the bifurcated reality of the repair.

```{r}
#| label: fig-multimodal-pert
#| fig-width: 8
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Example of a Bimodal Mixture Distribution. The mathematical 'average' time falls in a dead zone, highlighting why stochastic modeling is required for accurate labor planning."

library(ggplot2)
library(dplyr)

# Custom PERT density function to avoid extra package dependencies
dpert <- function(x, a, m, b) {
  alpha <- 1 + 4 * (m - a) / (b - a)
  beta <- 1 + 4 * (b - m) / (b - a)
  stats::dbeta((x - a) / (b - a), alpha, beta) / (b - a)
}

# Define the two states for the plumbing example
x <- seq(0, 120, length.out = 500)
state1 <- dpert(x, a = 5, m = 15, b = 30) * 0.70  # Tech adjustment (70% probability)
state2 <- dpert(x, a = 45, m = 70, b = 110) * 0.30 # Plumber PVC fix (30% probability)
mixture <- state1 + state2

df <- data.frame(Time = x, Density = mixture)
mean_time <- sum(df$Time * df$Density) / sum(df$Density)

ggplot(df, aes(x = Time, y = Density)) +
  geom_area(fill = "#4e79a7", alpha = 0.6) +
  geom_line(color = "#2c4663", linewidth = 1) +
  geom_vline(xintercept = mean_time, color = "#e15759", linetype = "dashed", linewidth = 1) +
  annotate("text", x = 15, y = max(df$Density) * 0.6, label = "State 1:\nTech Adjustment", size = 4) +
  annotate("text", x = 75, y = max(df$Density) * 0.4, label = "State 2:\nPlumber PVC Fix", size = 4) +
  annotate("text", x = mean_time + 2, y = max(df$Density) * 0.9, label = "Mean Time\n(Danger Zone)", color = "#e15759", hjust = 0, size = 4) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Mixture Distribution of Execution Time",
    x = "Execution Time (Minutes)",
    y = "Probability Density"
  ) +
  theme(panel.grid.minor = element_blank())
```

By mapping these mixture distributions across the entire property, we build a comprehensive dictionary of constraints. The complete operational matrix—detailing baseline defect rates, multi-state PERT parameters, material costs, and prerequisite dependencies—is defined in @tbl-tasks-config.

This master configuration codifies ground-level realities directly into the simulation engine. For example, it explicitly maps the physical handoffs between internal technicians adjusting standard sink drains and licensed plumbers replacing downstream PVC P-traps via vendor flags and scheduling lags. It also inherently respects the durability of the physical asset, mathematically ensuring that robust fixtures, such as existing upper kitchen cabinets, are not scheduled for costly full replacements unless critically broken.

Using the interactive table below, expanding the arrow next to any specific task reveals the exact time, cost, and probability parameters driving its Monte Carlo realizations.


```{r}
#| label: tbl-tasks-config
#| tbl-cap: "NSPIRE Task Modalities and Execution Constraints. Task categories map to core operational domains: FS (Fire Safety), EL (Electrical), PL (Plumbing), VA (Ventilation & Appliances), DW (Doors, Windows & Millwork), and SU (Surfaces). Clicking on the row arrow will display PERT, cost and decay parameters."
#| echo: false
#| warning: false
#| message: false

# Load the required libraries
library(DT)
library(readr)

# Read the CSV
tasks_df <- read_csv("../inst/extdata/tasks_config.csv")

# Generate and print the interactive table
datatable(
  tasks_df,
  extensions = c('FixedColumns', 'Responsive'),
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    fixedColumns = list(leftColumns = 5), 
    autoWidth = TRUE,
    columnDefs = list(
      list(className = 'dt-center', targets = "_all")
    )
  ),
  class = 'cell-border stripe hover'
)
```


## 3.4. Post-Audit Asset Decay

A final, critical parameter embedded within the master configuration (@tbl-tasks-config) is the baseline decay rate (p_decay). In residential properties, a resolved defect does not stay resolved indefinitely. Once a repair is completed and audited at time $t_{end}$, the asset immediately re-enters a state of stochastic wear-and-tear.

We model this defect re-emergence as a Bernoulli process over the waiting period $\delta_{decay}$ (the time gap between the completion of the repair and the arrival of the HUD inspector). Given a baseline daily probability of failure $\lambda_{i}$ for task $i$, the probability $P_{redefect}$ that a newly resolved task breaks again before the inspection is:
$$P_{redefect} = 1 - (1 - \lambda_{i,u}^*)^{\delta_{decay}}$$
Crucially, this decay is not uniform across the property. The effective daily probability of failure $\lambda_{i,u}^*$ is coupled with the unit's latent state $\phi_u$. This mathematical multiplier actively accelerates the re-emergence of defects in high-impact ("dirty") units:$$\lambda_{i,u}^* = \lambda_{i} \cdot \phi_u$$This temporal decay creates a severe operational trade-off. Finishing preparations too early maximizes the decay window ($\delta_{decay}$), inflating the probability that critical life-safety items (like a smoke detector or GFCI) will randomly fail before the inspector arrives. Conversely, starting preparations too late constricts the repair window, ensuring a labor backlog.

Beyond stochastic time estimates and decay rates, the master configuration table (@tbl-tasks-config) contains one more vital operational constraint: the prerequisites column. Real-world maintenance cannot be executed in a random order. To enforce the physical sequence of repairs, the simulation translates these prerequisites into a Directed Acyclic Graph (DAG), enforcing the rigid "order of operations" required to prep a unit.

# 4. Resource Mapping and Graph Topology (DAG)

In property maintenance, repairs cannot always happen simultaneously or in random order. For example, you cannot paint a wall before patching the drywall, and you cannot patch the drywall before fixing the plumbing leak behind it.

To model this operational reality, we organize maintenance tasks into a Directed Acyclic Graph (DAG). In simpler terms, a DAG is a strict, one-way flowchart that maps out the required "order of operations." Each task (mathematically called a vertex, $V_i$) requires an expected execution time $t_i$ and a material cost $m_i$.

```{mermaid}
%%| fig-cap: "A simple Directed Acyclic Graph (DAG) illustrating mandatory task sequences."
flowchart LR
    A[Fix Plumbing Leak] --> B[Patch Drywall]
    B --> C[Wait for Compound to Dry]
    C --> D[Paint Wall]
    
    style A fill:#f9f9f9,stroke:#333,stroke-width:2px
    style B fill:#f9f9f9,stroke:#333,stroke-width:2px
    style C fill:#ffe6cc,stroke:#d79b00,stroke-width:2px,stroke-dasharray: 5 5
    style D fill:#f9f9f9,stroke:#333,stroke-width:2px
```

When applied to the entire NSPIRE catalog, the complexity of property maintenance emerges. In @fig-dag-network below, the algorithm scales the size of each node by its mathematical out-degree: the number of downstream tasks that physically depend on it.

In property management terms, massive nodes represent bottlenecks. For example, EL-02 (the Breaker Box) acts as a massive prerequisite hub. If this root node is not cleared first, it initiates a cascading scheduling freeze, preventing technicians from executing subsequent tasks like GFCI replacements, exhaust fan repairs, or lighting installations.

```{r}
#| label: fig-dag-network
#| fig-cap: "The complete maintenance DAG. Node size indicates bottleneck severity. Hover over any node to view the full task description. Use the top control in the right corner to reset the view. Categories of tasks: FS (Fire Safety), EL (Electrical), PL (Plumbing), VA (Ventilation & Appliances), DW (Doors, Windows & Millwork), and SU (Surfaces)"
#| echo: false
#| warning: false
#| message: false

library(igraph)
library(visNetwork)

# 1. Load the DAG and Configuration
dag_csv <- "../inst/extdata/tasks_config.csv"
dag_obj <- create_inspection_dag(dag_csv)
g <- dag_obj@graph
tasks_df <- dag_obj@tasks_config

# 2. Build Interactive Nodes Dataframe
nodes <- data.frame(id = V(g)$name, label = V(g)$name)

# Match node IDs to config to extract categories and descriptions
idx <- match(nodes$id, tasks_df$task_id)
nodes$group <- tasks_df$category[idx]

# Create the tooltip (HTML formatted)
nodes$title <- paste0("<p><b>", tasks_df$task_id[idx], "</b><br>", 
                      tasks_df$description[idx], "</p>")

# Calculate out-degree for node sizing
out_deg <- degree(g, mode = "out")
nodes$value <- out_deg + 1 

# 3. Build Interactive Edges Dataframe
edges <- igraph::as_data_frame(g, what = "edges")

# 4. Render Interactive Graph
visNetwork(nodes, edges, width = "100%", height = "600px") |>
  visIgraphLayout(layout = "layout_with_fr", randomSeed = 42) |>
  visEdges(arrows = "to", color = list(color = "#a9a9a9"), smooth = FALSE) |>
  visNodes(
    font = list(size = 20, face = "sans", color = "black"),
    borderWidth = 1
  ) |>
  visOptions(
    highlightNearest = list(enabled = TRUE, degree = 1, hover = FALSE),
    nodesIdSelection = TRUE # Dropdown menu to search for specific nodes
  ) |>
  # Adds the navigation and reset controls
  visInteraction(navigationButtons = TRUE)
```


## 4.1. Edge Constraints (Operational Laws)

The arrows connecting these tasks are not just visual lines; they act as rigid operational laws within the simulation engine.

* **Sequential Precedence**: A directed edge $V_A \to V_B$ requires the start time of $B$ to be $\ge$ the end time of $A$.
  * Example: A technician cannot paint a wall before the drywall compound has fully dried.

* **Temporal Blocking**: Certain vertices impose a local resource freeze, locking a spatial node for a time interval $\Delta t$.
  * Example: If an external vendor reglazes a tub that physical zone is locked out for a mandatory cure time. The scheduling engine forces internal technicians to re-route to other tasks or units.

* **Mutually Exclusive Pruning**: If a higher-order vertex $V_{replace}$ is activated, redundant child vertices $V_{repair}$ are mathematically pruned ($t_{repair} \to 0$).
  * Example: If an upper kitchen cabinet is critically broken and flagged for full replacement, the system instantly cancels minor, redundant adjustments to its hinges or doors, preventing the budget from absorbing phantom labor hours.


# 5. Execution Capacity and Quality Control

Even with a perfectly sequenced DAG, execution is bound by strict physical limits: the time available to work, the reliance on external vendors, and the inevitability of human error.

## 5.1. Repair Capacity and Vendor Bottlenecks

The model assumes a maintenance repair window defined by $t_{start}$ (commencement of internal repairs) and $t_{end}$ (completion of repairs and audit). Given a standard baseline labor capacity $R_{base}$ (hours/unit/day), the total labor capacity $\text{Cap}$ available for the maintenance cycle is linearly proportional to the duration of this window:
$$\text{Cap} = R_{base} \cdot (t_{start} - t_{end})$$
Tasks requiring labor hours $T_{total} > \text{Cap}$ result in a scheduling backlog, forcing the system to abandon lower-severity items to prioritize high-penalty (LT/S) defects.

This capacity is further choked by tasks requiring external licensing. For example, while a technician handles a sink drain, repairing the downstream PVC P-trap requires a licensed plumber. When a node evaluates to this severe state, internal labor drops to zero ($t_{internal} \to 0$), but a schedule lag is injected based on a Uniform distribution: $L \sim \text{Uniform}(l_{min}, l_{max})$ days. Furthermore, a mandatory spatial lockout is triggered (e.g., a 24-hour cure time for glue or silicone), forcing the DAG to re-route internal technicians and accelerating the exhaustion of the remaining timeline.

## 5.2. Human Error and Audit Efficacy

Execution of a task does not guarantee absolute resolution. A task has a baseline physical probability ($p_{err}$) of being missed or improperly repaired. We quantify quality control using a formal internal audit efficacy parameter, $\alpha_\text{audit} \in [0, 1]$, representing the probability that the pre-inspection internal team detects and successfully mitigates a surviving defect.

* **Serial Execution (Single Technician)**: The probability that a defect survives the initial repair attempt and remains undetected by the audit phase is:
$$P(\text{Survive}) = (1 - \alpha_\text{audit}) \cdot p_{err}$$

* **Parallel Execution (Dual Technicians)**: Applying the "Four Eyes" principle, independent cross-checking drastically reduces error. The joint probability of both technicians missing the defect becomes:$$P(\text{Survive}) = (1 - \alpha_\text{audit}) \cdot p_{err}^2$$

# 6. Tenant Heterogeneity and Predictive Risk

Asset deterioration is not uniform across a property. Baseline probabilities assume a standard deterioration rate, but operational reality dictates that damage clusters around specific tenant behaviors.

## 6.1. The Bimodal Risk of Tenant Behavior

We model unit "cleanliness" as a bimodal latent state $\phi_u$ for each unit $u$, representing the inherent maintenance burden generated by the tenant. The tenant impact multiplier $\phi_u$ is drawn from a mixture distribution:
$$\phi_u \sim \begin{cases} \phi_{clean}, & \text{with probability } (1 - \pi) \\ \phi_{dirty}, & \text{with probability } \pi \end{cases}$$
Where $\pi$ is the estimated proportion of high-impact units in the property.

## 6.2. The Danger of Silence (Latent Covariates)

A lack of historical tenant reporting often masks severe deferred maintenanc. To combat this, we use historical service request volume $W_u$ as a proxy to infer the probability that a unit belongs to the high-impact "dirty" latent state ($\phi_{dirty}$).

We define the unit-specific probability $\pi_u$ as a logistic function of service requests $W_u$:
$$\pi_u = P(\phi_u = \phi_{dirty} \vert{} W_u) = \text{logit}^{-1}(\alpha - \beta W_u)$$
Where $\alpha$ and $\beta$ are calibrated tuning parameters. For units with extremely low reporting ($W_u \to 0$), the prior probability of being in the "dirty" state dramatically increases, signaling a high likelihood of latent defects.

During simulation initialization, this unit-specific $\pi_u$ dynamically replaces the global mixture probability. Each unit's tenant factor $\phi_u$ is sampled conditioned directly on its specific reporting history. This transforms the simulation from a static property model into a predictive prioritization tool. By mapping historical silence to bimodal risk, the engine directs the maintenance DAG to audit "quiet" units before they enter the randomized HUD sampling window.

# 7. Risk Analysis

Moving beyond deterministic score projections, this model quantifies operational risk by treating property performance as a stochastic process. By leveraging Monte Carlo simulation, we assess not only the expected outcome but the exact mathematical probability of regulatory non-compliance.

7.1. Quantification of Failure Risk

We define the failure event as $S_{eff} < 60$. Due to the non-linear interaction of random unit sampling, defect emergence probabilities, and repair quality, the effective inspection score $S_{eff}$ is a random variable. We estimate the probability of failure ($P_{fail}$) by simulating $K = 1,000$ independent inspection realizations:
$$P_{fail} = \frac{1}{K} \sum_{i=1}^{K} I(S_{eff,i} < 60)$$
where $I(\cdot)$ is the indicator function. This allows us to map the property's risk profile against the HUD enforcement threshold, providing a granular view of the likelihood of catastrophic failure that a mean-based score effectively masks.

## 7.2. Quantifying Volatility 

The Tail Drop MetricStandard deviation ($\sigma$) and the Coefficient of Variation ($CV = \sigma/\mu$) provide a global measure of dispersion. However, in NSPIRE compliance, the right tail (scores $> 100$) is truncated, and the distribution is often skewed. The primary operational risk resides in the left tail—the possibility of a "bad day" where HUD sampling captures a concentrated cluster of high-severity defects.To quantify this downside volatility, we define the Tail Drop Metric ($\tau$):

$$\tau = \frac{\text{med} - P_{05}}{\text{med}}$${#eq-tau}

Where:

* $\text{med}$ is the median score.$P_{05}$ is the 5th percentile of the simulated scores (the "Value at Risk").
* $\tau$ represents the percentage degradation from expected performance to the worst-case scenario. 

A property may exhibit a high median score while maintaining a high $\tau$, identifying it as a "lottery" asset—stable on average, but highly susceptible to non-linear drops in score due to the interaction of unit sampling variance and tenant-driven decay.

## 7.3. Sensitivity Mapping

The Tail Drop Metric $\tau$ is a systemic output of our model, functionally defined by the interaction of our control and stochastic parameters:

$$\tau = f(n(N), \alpha_\text{audit}, \pi, \phi_{dirty})$$

This sensitivity function captures how variance in the HUD sampling tier $n(N)$, human error in pre-inspection audit ($\alpha_\text{audit}$), and the underlying distribution of tenant impact ($\pi, \phi_{dirty}$) propagate into downside performance risk. The following numerical investigations isolate these partial derivatives to determine which operational parameters dominate the variance in the $P_{05}$ left tail.


# 8. Operational Reality: The Transit Friction Tax

Theoretical models that treat technicians as perfect, frictionless robots are likely to provide unrealistic time estimates. Standard algorithmic approaches would calculate only pure "wrench time"—the isolated minutes required to physically execute a repair. This fundamentally ignores the physical and human constraints of moving labor through a segmented residential environment.

To accurately model ground-level execution, the simulation injects "transit friction" into the schedule:

* **Intra-Unit Friction (~5 minutes)**: The time required to clean up, swap tools, and transition between tasks within the same apartment. Even minor operational handoffs—such as a technician finishing a standard sink drain adjustment before a plumber takes over the downstream PVC—introduce rigid micro-delays.
* **Inter-Unit Friction (~15 minutes)**: The time required to pack up a tool bag, navigate the property, unlock a new door, and assess the next cluster of defects. Crucially, this also accounts for basic human necessities: taking a water break, using the restroom, or pausing to answer a colleague's question.

To quantify the macroeconomic impact of these human constraints, the simulation executed 10,000 paired scenarios (comparing pure wrench time against friction-adjusted time) for a standard 50-unit property.

```{r}
#| label: fig-transit-friction
#| fig-width: 9
#| fig-height: 6
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Transit Friction: including short breaks after a unit task (~5min) and slightly longer ones (~15min) when swiching to a different unit induces a 34.2% tax time."

plot_data <- readRDS("../results/transit_friction/friction_plot.rds")
paired_data <- readRDS("../results/transit_friction/friction_distribution_data.rds")
mean_ideal <- mean(paired_data$Ideal)
mean_real <- mean(paired_data$Realistic)
tax_pct <- ((mean_real - mean_ideal) / mean_ideal) * 100
p <- ggplot(plot_data, aes(x = Total_Hours, fill = Scenario, color = Scenario)) +
  geom_density(alpha = 0.6, linewidth = 1) +
  scale_fill_manual(values = c("Ideal" = "#4e79a7", "Realistic" = "#e15759"),
                    labels = c("Theoretical Wrench Time", "Realistic Time (with Friction)")) +
  scale_color_manual(values = c("Ideal" = "#2c4663", "Realistic" = "#8c2d2e"), guide = "none") +
  theme_minimal(base_size = 14) +
  labs(
    title = "The Operational Reality Gap",
    subtitle = sprintf("Transit friction adds an average %.1f%% 'tax' to total required labor hours.", tax_pct),
    x = "Total Labor Hours Required",
    y = "Density (Probability)"
  ) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  )

print(p)
```

## Conclusions

**The 34% Labor Tax**: Transit friction is not a symptom of an inefficient workforce; it is a structural penalty dictated by physics and human reality. The compounding effect of minor 5-minute and 15-minute delays generates an unavoidable 34.2% expansion in total required labor hours.

**Budgetary Blind Spot**: If management calculates staffing requirements, prep timelines, or overtime budgets based solely on idealized task lists, the property is instantly under-resourced by a third. Optimizing execution requires scheduling algorithms that prioritize spatial clustering, keeping technicians in a single unit as long as possible to suppress this hidden tax.

# 9. Parameter Sensitivity Analysis

Sensitivity analysis studies how variations in control parameters drive system outcomes, such as expected score, labor costs, and downside risk ($P_{fail}$ and $\tau$). By studying these variables across parameter ranges, one can identify parameter critical ranges, non-linear thresholds and propose optimal resource allocations within the NSPIRE framework.

## 9.1. Audit Catch Rate ($\alpha_\text{audit}$)

The internal pre-inspection audit is modeled by $\alpha_\text{audit}$, the probability that a defect is identified and resolved prior to the HUD inspection. To quantify the impact of audit efficacy on compliance risk, $\alpha_\text{audit}$ was varied from $0.05$ (negligent) to $0.95$ (highly rigorous), while executing $K=1,000$ Monte Carlo iterations per step. @fig-audit-ridges provides a visual summary of the resulting score distributions (horizontal axis with a red line for the failing score of 60) as a function of $\alpha_\text{audit}$ (vertical axis).

```{r}
#| label: fig-audit-ridges
#| fig-width: 9
#| fig-height: 6
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Risk Transition: Audit Rate vs. Score Distribution."

sweep_data <- load_sensitivity_iterations(here("results", "sensitivity", "audit_rate"))

# Generate the plot
plot_sensitivity_ridges(
  sensitivity_df = sweep_data,
  param_label = "Audit Catch Rate (Efficacy)",
  title = "Audit Rate vs. Score Distribution"
)
```

Failure risk is almost completely eliminated when $\alpha_\text{audit} \geq 0.9$. @fig-distribution-slice further explains the transition from high risk to low risk as $\alpha_\text{audit}$ increases from 0.55 to 0.85. The failure risk is best estimated with the proportion of simulations (among $K=1,000$) that yield a score below 60 (red vertical line). Visually, this corresponds to the area under the density curve (integral) to the left of the critical red line. 


```{r} 
#| label: fig-distribution-slice
#| fig.width: 12 
#| fig.height: 6
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Examples of Score Distribution: $\\alpha_\\text{audit} = 0.55$ (left) and $\\alpha_\\text{audit} = 0.85$ (right)."

# 1. Define the scenarios
scenarios <- c(0.55, 0.85)
sum_df <- readRDS(here("results", "sensitivity", "audit_rate", "summary.rds"))
pal <- scales::viridis_pal(option = "D")(100) # Pre-generate palette

plots <- lapply(scenarios, function(rate) {
  # 1. Cast both to numeric to ensure they match
  median_val <- sum_df$score_median[as.numeric(sum_df$param_val) == as.numeric(rate)]
  
  # 2. Pick color based on median score (0-100 mapped to 1-100 index)
  idx <- max(1, min(100, round(median_val)))
  
  plot_score_distribution(
    input = here("results", "sensitivity", "audit_rate", paste0("score_density_", rate, ".rds")),
    title = paste("Audit Catch Rate:", rate * 100, "%"),
    color_hex = pal[idx]
  )
})

plots[[1]] | plots[[2]]

```

Estimating both failure risks (score less than 30 or less than 60) at all values of $\alpha_\text{audit}$ yields @fig-failure-audit. The data demonstrates that critical failure risk (score < 30) is only marginalized when audit catch rates exceed 0.75, while full protection against standard failure (score < 60) requires efficacy above 0.90.

```{r}
#| label: fig-failure-audit
#| fig.width: 6
#| fig.height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Failure probabilities at scores 30 and 60 as a function of audit catch rate $\\alpha_\\text{audit}$."

sum_df <- readRDS(here("results", "sensitivity", "audit_rate", "summary.rds"))

plot_sensitivity_multi(
  sum_df, 
  y_vars = c("fail_risk_60", "critical_fail_risk_30"),
  param_label = "Audit Catch Rate",
  y_label = "Probability",
  title = "Impact of Audit Efficacy on Regulatory Failure"
)
```

Besides failure probabilities, another useful statistic is the tail-drop metric $\tau$ @eq-tau. To understand its sensitivity to $\alpha_\text{audit}$, we must first examine how the foundational score statistics evolve across the parameter range. These plots are displayed in @fig-dist-stats-audit.

```{r}
#| label: fig-dist-stats-audit
#| fig.width: 6
#| fig.height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Score statistics as a function of audit catch rate $\\alpha_\\text{audit}$."

sum_df <- readRDS(here("results", "sensitivity", "audit_rate", "summary.rds"))

plot_sensitivity_multi(
  sum_df, 
  y_vars = c("score_p05", "score_median", "score_sd"),
  param_label = "Audit Catch Rate",
  y_label = "Score",
  title = "Impact of Audit Efficacy on Score Statistics"
)
```

As $\alpha_\text{audit}$ increases beyond 0.25, the score median starts rising linearly. Later, as $\alpha_\text{audit}$ gets above 0.45, the 5th percentile score $P_{05}$ also rises linearly but with a stiffer slope. This yields decreasing differences between median and 5th percentile. One way to visualize this is to plot the standard deviation of the score (a proxy measure for difference between median and 5th percentile) as a function of $\alpha_\text{audit}$. These behaviors of the different statistics yield the profile of $\tau$ @eq-tau as a function of $\alpha_\text{audit}$ displayed in @fig-tau-audit.

```{r}
#| label: fig-tau-audit
#| fig.width: 6
#| fig.height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Tail drop metric $\\tau$ as a function of audit catch rate $\\alpha_\\text{audit}$."

sum_df <- readRDS(here("results", "sensitivity", "audit_rate", "summary.rds"))

plot_sensitivity_multi(
  sum_df, 
  y_vars = c("tail_drop_median"),
  param_label = "Audit Catch Rate",
  y_label = "Tail Drop Metric",
  title = "Impact of Audit Efficacy on Tail Drop Metric."
)
```

### Conclusions

* **Operational Futility Threshold ($\alpha_\text{audit} \leq 0.25$)**: Below ~25%
efficacy, quality assurance efforts are mathematically invisible to the final metric due to the absolute scoring floor (0).

* **Variance Squeeze ($0.45 < \alpha_\text{audit} < 0.90$)**: As the catch rate rises above 0.45, the 5th percentile worst-case outcome ($P_{05}$) improves at a faster linear rate than the median. This compresses the score variance and significantly reduces the Tail Drop metric ($\tau$).

* **Safety Baseline ($\alpha_\text{audit} \geq 0.90$)**: To confidently eliminate standard regulatory failure risk (score < 60), the internal audit process must successfully intercept at least 90% of all pre-existing defects.



## 9.2. The Time Crunch: Notice of Inspection ($t_\text{start}$)

While a robust audit catch rate ($\alpha_\text{audit}$) is critical for passing an NSPIRE inspection, it assumes the maintenance team has the physical time to execute the repairs. HUD inspection notices can vary drastically, sometimes providing as little as 14 to 30 days of warning.

To model this operational squeeze, the preparation window ($t_\text{start}$) was swept from 36 days down to 4 days. For this specific scenario, the internal audit catch rate was fixed at a highly effective 85% ($\alpha_\text{audit} = 0.85$), and maximum allowable overtime was expanded to 4 hours per unit per day. Crucially, the simulation incorporates context-aware transit friction: technicians incur small delays (1–5 minutes) when moving between tasks in the same unit, but significant delays (15–30 minutes) when packing up and accessing a new unit.

As the preparation window shrinks, the volume of identified defects remains constant, but the time available to clear them—and the time available to walk between them—compresses. 

```{r}
#| label: fig-time-crunch-ridges
#| fig-width: 9
#| fig-height: 6
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Score Distribution Evolution under time compression. Ridges are shown at 4-day intervals for visual clarity."

sweep_data <- load_sensitivity_iterations(here("results", "sensitivity", "t_start"))

# Subset to avoid visual clutter (e.g., plotting every 4 days)
subset_days <- seq(4, 36, by = 3)
sweep_data_sub <- sweep_data[as.numeric(sweep_data$param_val) %in% subset_days, ]

# Generate the plot
plot_sensitivity_ridges(
  sensitivity_df = sweep_data_sub,
  param_label = "Days of Notice (t_start)",
  title = "Time Crunch vs. Score Distribution"
)
```

As visualised in @fig-time-crunch-ridges, the score distribution remains stable when $t_\text{start} \geq 28$. However, as time compresses, the distribution stretches heavily to the left before completely fracturing and collapsing into the failure zone.

The initial impact is purely financial, as visualized in @fig-time-crunch-labor.

```{r}
#| label: fig-time-crunch-labor
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Mean total labor cost and overtime cost as the preparation window ($t_\\text{start}$) changes. Note the massive overtime spike at Day 22 followed by the sharp drop in total costs as the timeline collapses."

sum_df_time <- readRDS(here("results", "sensitivity", "t_start", "master_time_crunch_study.rds"))

plot_sensitivity_multi(
  sum_df_time, 
  y_vars = c("total_cost_mean", "overtime_mean"),
  param_label = "Days of Notice (t_start)",
  y_label = "Cost ($)",
  title = "Financial Impact of Compressed Prep Windows"
)
```

At first glance, the left side of @fig-time-crunch-labor appears counter-intuitive: as the preparation window drops below 20 days, the total expected cost plunges dramatically. This is not an efficiency gain; it is the mathematical manifestation of deferred maintenance.

Between Day 30 and Day 22, the workforce heavily leverages their overtime allowance to keep pace with the workload. At Day 22, the system hits peak labor output—the team is operating at absolute maximum capacity, driving overtime spend to its highest point. However, as the window shrinks below 20 days, the absolute physical capacity of the workforce is exhausted. Hundreds of items are forced into the "backlog." Because backlogged items are abandoned, they incur no labor and no material costs, creating a "false economy" of savings.

@fig-time-crunch-backlog illustrates the severe regulatory consequences of this capacity exhaustion.

```{r}
#| label: fig-time-crunch-backlog
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Probability of generating a task backlog and the resulting regulatory failure risk ($<60$) under compressed timelines."

plot_sensitivity_multi(
  sum_df_time, 
  y_vars = c("backlog_prob", "fail_risk_60", "critical_fail_risk_30"),
  param_label = "Days of Notice (t_start)",
  y_label = "Probability",
  title = "The Backlog Cliff: Capacity Exhaustion"
)
```

Finally, we map the Tail Drop metric ($\tau$) as a function of the preparation window. This demonstrates how temporal limits affect the volatility of outcomes.

```{r}
#| label: fig-time-crunch-tau
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Tail Drop Metric $\\tau$ as a function of the preparation window. High $\\tau$ indicates massive outcome volatility."

plot_sensitivity_multi(
  sum_df_time, 
  y_vars = c("tail_drop_median"),
  param_label = "Days of Notice (t_start)",
  y_label = "Tail Drop Metric",
  title = "Impact of Prep Window on Tail Drop Metric"
)
```

### Conclusions

**Distributed Defect Tax ($t_\text{start} \geq 28$)**: Because repairing widely distributed defects incurs a heavy transit-time penalty, a 28-day notice is the edge of the safe operational limit for this property. At 28 days, the property maintains a near 0% failure risk with minimal overtime expenditure.

**Maximum Effort Peak ($t_\text{start} \approx 22$)**: Day 22 represents the system's breaking point. Overtime spend reaches its absolute maximum as the team attempts to outwork the shrinking timeline. Despite spending heavily on premium labor, the backlog probability reaches ~50%, and the property begins to exhibit a measurable risk of regulatory failure. The team is throwing maximum capital at the problem but is still losing ground.

**Operational Collapse ($t_\text{start} \leq 15$)**: A standard 14-day HUD notice is mathematically unfeasible for a property in this state and with 5 technicians. By Day 15, the probability of backlogging tasks reaches 100%, and the property's risk of failure exceeds 85%. The simulation proves that sudden drops in preparation expenditure in the days leading up to an inspection are a lagging indicator of capacity exhaustion, representing a forced trade-off of immediate OpEx savings for guaranteed regulatory failure.


## 9.3. Prescriptive Analytics: The Staffing Curve

Having established in Section 9.2 that a compressed timeline guarantees regulatory failure due to capacity exhaustion, we can invert the simulation to perform prescriptive analytics. By decoupling the labor constraints from the QA blind spot, we can query the engine: Given a specific preparation window, what is the exact minimum headcount required to guarantee a 99% probability of clearing the known backlog?

Using a binary search algorithm wrapped around the Monte Carlo engine (with $K=1,000$ iterations per step), we mapped the required workforce capacity as the timeline collapses. We targeted a strict operational threshold ($P(\text{Backlog} > 0) \le 0.01$) to ensure the recommended staffing levels are mathematically robust.

```{r}
#| label: fig-staffing-curve
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Prescriptive Staffing Curve: Minimum technicians required to guarantee $P(\\text{Backlog}) \\le 0.01$ as the inspection window compresses."

# Load the high-resolution staffing curve data
# Assuming it was saved via: saveRDS(staffing_df, here("results", "sensitivity", "t_start", "staffing_curve.rds"))
staffing_df <- readRDS(here::here("results", "sensitivity", "t_start", "staffing_curve.rds"))

ggplot2::ggplot(staffing_df, ggplot2::aes(x = t_start, y = required_techs)) +
  ggplot2::geom_line(color = "#8b0000", linewidth = 1.2, direction = "vh") +
  ggplot2::geom_point(shape = 21, fill = "white", color = "#8b0000", size = 2, stroke = 1.2) +
  ggplot2::scale_x_reverse() + # Reverse X-axis so time counts down to the right
  ggplot2::scale_y_continuous(breaks = scales::pretty_breaks(n = 6)) +
  ggplot2::theme_minimal(base_size = 14) +
  ggplot2::labs(
    title = "Prescriptive Staffing Curve",
    x = "Days of Notice (t_start)",
    y = "Required Technicians"
  ) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    plot.title = ggplot2::element_text(face = "bold")
  )
```

The resulting curve is highly non-linear. With a generous notice period, the baseline crew of technicians is perfectly adequate to absorb the transit friction and complete the repairs. However, dropping below the 15-day threshold triggers an exponential surge in labor demand. To prep this property in under a week, a property manager cannot simply authorize overtime; they must instantly deploy a massive, coordinated swarm of technicians.

### Conclusions

This prescriptive tool translates abstract regulatory timelines directly into actionable HR and budgeting directives, allowing management to definitively calculate the cost of a delayed response.



## 9.4. Post-Prep Decay: The "Too Early" Penalty ($t_\text{end}$)

In Section 9.2, we established that compressing the preparation timeline forces a property into a backlog cliff due to labor limits. A common operational response to this risk is to overcompensate by completing all pre-inspection repairs well in advance.

However, residential properties are dynamic, entropic environments subject to continuous wear-and-tear. The time gap between the completion of repairs ($t_\text{end}$) and the actual day of the HUD inspection exposes the property to a natural re-accumulation of defects.

To quantify this temporal decay risk, we swept $t_\text{end}$ from 0 days (repairs finish the morning of the inspection) up to 35 days. To isolate the effect of decay from capacity constraints, the starting notice ($t_\text{start}$) was locked at 45 days, ensuring the team had abundant capacity and zero backlog risk.

@fig-decay-score illustrates the erosion of the property's safety margin as the time gap widens.

```{r}
#| label: fig-decay-score
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Erosion of the Median and 5th Percentile ($P_{05}$) scores as the gap between prep completion and the inspection date widens."

sum_df_decay <- readRDS(here::here("results", "sensitivity", "t_end", "master_decay_study.rds"))

plot_sensitivity_multi(
  sum_df_decay, 
  y_vars = c("score_median", "score_p05"),
  param_label = "Days Finished Before Inspection (t_end)",
  y_label = "Score",
  title = "Impact of Post-Prep Decay on Score Confidence"
)
```

As visualized in @fig-decay-score, the score does not degrade at a steady, linear rate. Instead, the property exhibits strong resilience for the first three weeks, maintaining stable median and $P_{05}$ scores up to $t_\text{end} \approx 25$. However, beyond 25 days, the score plunges rapidly. This indicates that temporal decay acts as a threshold effect: accumulating minor wear-and-tear is initially absorbable, but as time passes, the compounding probability of a critical life-safety system (e.g., a smoke detector or GFCI outlet) failing at random reaches a critical mass, shattering the score.

@fig-decay-risk visualizes how this sudden score erosion translates into concrete regulatory failure.

```{r}
#| label: fig-decay-risk
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "The compounding probability of regulatory failure caused by natural property decay during the post-prep waiting period."

plot_sensitivity_multi(
  sum_df_decay, 
  y_vars = c("fail_risk_60", "critical_fail_risk_30"),
  param_label = "Days Finished Before Inspection (t_end)",
  y_label = "Probability",
  title = "Failure Risk Amplification over Time"
)
```

### Conclusions:

**Decay Cliff**: Completing preparations too early introduces severe mathematical risk, manifesting as a "cliff" rather than a steady slope. The data demonstrates that a property remains relatively safe if it sits idle for up to three weeks. However, a property that is highly likely to pass at $t_\text{end} = 0$ exhibits an exponential spike in failure risk if left idle beyond 25 days, reaching near-certain failure by day 35.

**"Goldilocks" Prep Window**: Combined with the capacity insights from Section 9.2, these results define a rigid operational window. A maintenance team must start late enough to avoid an excessive decay penalty ($t_\text{end} \le 25$ days) but early enough to avoid capacity exhaustion ($t_\text{start}$ constraint). Maximizing NSPIRE performance requires precise "just-in-time" repair scheduling.



## 9.5. Spatial Heterogeneity and the "Lottery Effect"

NSPIRE scores are heavily influenced by the spatial distribution of defects. HUD's sampling protocol selects a subset of units to represent the entire property (e.g., sampling 20 units in a 50-unit building). While this works mathematically for uniformly maintained properties, it introduces massive variance—the "Lottery Effect"—in properties with severe tenant heterogeneity.

To quantify this, we varied the clean_proportion parameter from 1.00 (homogeneous: 100% normal tenants) down to 0.70. To isolate the sampling risk and ensure we were not artificially degrading the building's overall quality, a dynamic balancing equation was used. We enforced a strict "Conservation of Defects" by fixing the global expected severity multiplier at 1.0. 

Let $p$ be the proportion of clean units and $R$ be the severity ratio of a destructive unit (set to $20.0$). The expected defect multipliers for clean units ($\mu_c$) and destructive units ($\mu_d$) must satisfy:

$$p \mu_c + (1 - p) \mu_d = 1.0$$

By substituting $\mu_d = R \mu_c$, we dynamically calculate the baseline multiplier for the clean units for any given tenant mixture:

$$\mu_c = \frac{1}{p + R(1 - p)}$$

Consequently, as destructive tenants are introduced and generate severe defect clusters, standard units become proportionally cleaner to offset them.

To explicitly measure how this sampling variance impacts regulatory risk, we apply the Tail Drop Median metric, $\tau$ (@eq-tau), which isolates the expansion of the left tail relative to the median score. @fig-tail-risk illustrates how this tail risk behaves as a function of the tenant mix.

```{r}
#| echo: false
#| message: false
#| warning: false
library(dplyr)

# 1. Load the raw ridges data
ridges_df <- readRDS(here::here("results", "sensitivity", "tenant_mix", "master_tenant_ridges.rds"))

# 2. Extract median, P05, and calculate tau
tau_df <- ridges_df %>%
  group_by(param_val) %>%
  summarize(
    median_score = median(score),
    P05 = quantile(score, 0.05),
    .groups = "drop"
  ) %>%
  mutate(tail_drop_median = (median_score - P05) / median_score)
```

```{r}
#| label: fig-tail-risk
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Tail Risk Expansion. As the building deviates from perfect homogeneity, the left tail of the score distribution detaches from the median, maximizing the statistical risk of a catastrophic inspection."
plot_sensitivity_multi(
  tau_df,
  y_vars = "tail_drop_median",
  param_label = "Proportion of Clean Tenants",
  y_label = "Tail Drop Median (\u03c4)",
  title = "Tail Risk Expansion (\u03c4) vs. Tenant Heterogeneity"
)
```
Reading from right to left, as homogeneity breaks down, $\tau$ sharply increases, peaking near $p = 0.87$. This visualizes the core of the Lottery Effect: while the median score of the building remains relatively stable, the 5th percentile plummets. To understand the physical mechanisms driving this expansion—and its subsequent contraction at extreme heterogeneity—we must look at the physical defects themselves.


@fig-tenant-physics visualizes the physical state of the building prior to HUD scoring.

```{r}
master_tenant_study <- readRDS(here::here("results", "sensitivity", "tenant_mix", "master_tenant_study.rds"))
```

```{r}
#| label: fig-tenant-physics
#| layout-ncol: 2
#| fig-width: 5
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "The Physical Reality of Heterogeneity. As the building becomes more homogeneous (moving left to right), the spatial clustering of defects collapses (Plot B). The U-shape in total defects (Plot A) reveals the physical capacity limits of individual units."
#| fig-subcap: 
#|   - "Conservation of Defects"
#|   - "Spatial Clustering Spike"

# Plot A: Total Defects
plot_sensitivity_metric(
  master_tenant_study,
  y_var = "avg_total_defects",
  param_label = "Proportion of Clean Tenants",
  y_label = "Total Building Defects",
  title = "A: Conservation of Defects"
)

# Plot B: Clustering
plot_sensitivity_metric(
  master_tenant_study,
  y_var = "avg_clustering_sd",
  param_label = "Proportion of Clean Tenants",
  y_label = "Standard Deviation of Unit Defect Count",
  title = "B: Spatial Clustering Spike"
)
```

The data reveals a statistical paradox driven by physical constraints (Plot A). As the proportion of clean tenants increases (left to right), the balancing equation forces a shrinking number of destructive units to carry the mathematical burden. These few units quickly hit a physical ceiling—a single apartment can only harbor so many broken items. Consequently, excess theoretical defects are truncated, artificially dropping the building's total defect count. 

Once $p$ pushes past $0.94$, the burden shifts back to the clean units, which can easily absorb the expected defects without truncation, allowing the total volume to rapidly rebound to its theoretical expectation. 

Crucially, the SD of defect count per unit (Plot B) peaks before collapsing. This peak represents the maximum disparity between clean and destructive units. Beyond this threshold, building-wide variance plummets due to two intertwined factors: a shrinking population of destructive units to act as outliers, and physical item caps that prevent those remaining units from absorbing extreme defect multipliers.

@fig-tenant-risk demonstrates how this physical clustering translates into regulatory failure.

```{r}
#| label: fig-tenant-risk
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "The 'Lottery Effect' on Regulatory Risk. Despite having fewer overall defects, buildings with highly concentrated tenant damage experience a massive spike in inspection failure probability."

plot_sensitivity_multi(
  master_tenant_study, 
  y_vars = c("fail_risk_60", "critical_fail_risk_30"),
  param_label = "Proportion of Clean Tenants",
  y_label = "Probability of Failure",
  title = "Failure Risk vs. Tenant Heterogeneity"
)
```

Despite the heterogeneous building containing fewer total defects than the homogeneous baseline, its probability of regulatory failure more than triples (from 5% to 15%).

### Conclusions:

**The Lottery Effect is Real**: Under the NSPIRE scoring protocol, you cannot "average out" a destructive tenant. Because HUD's divisor is based on the sample size, stepping into just one or two heavily concentrated units will instantly drag an otherwise pristine property below a passing score.

**Targeted Pre-Inspections**: A property with fewer total defects but high clustering is statistically more dangerous than a uniformly degraded property. If management is aware of specific high-impact tenants, those specific units must be audited and repaired at a 100% catch rate prior to the inspection, as they pose a disproportionate mathematical threat.



<style>
.floating-copyright {
  position: fixed;
  bottom: 15px;
  left: 15px;
  font-size: 0.85em;
  color: #777;
  background-color: rgba(255, 255, 255, 0.85);
  padding: 5px 10px;
  border-radius: 5px;
  z-index: 1000;
  pointer-events: none; 
}
</style>

<div class="floating-copyright">
  &copy; 2026 Joël R. Pradines.<br> All rights reserved.
</div>